### 2025-01-07 paper.tex
.

\documentclass{article}
\usepackage[utf8]{inputenc}

\title{Compute Servers for Teaching Big Data}
\author{Clark Fitzgerald}
\begin{document}

\maketitle

\begin{abstract}

I’ve taught an upper division statistics ``Big Data'' course at UC Davis and CSU Sacramento and have gained some perspective on what works well for this course.
My goals with the course are to get students comfortable with using remote machines and to do realistic analysis of data sets that don’t fit in memory, typically on the order of 100GB or so.

I’ve tried what feels like an excessive number of ways to run the server for the students to get the `server experience':
Campus supported Linux cluster (SLURM)
AWS EC2 - student accounts and Jupyter Notebooks
Google colab
NSF Jetstream cloud (Access)
Physical server in the corner of my office (current solution!)

\end{abstract}

\end{document}

### 2025-01-07 paper.tex
.

@@
 
+The class is called `Big Data'.
+For this term to have any meaning, we need to define what we mean by `big'.
+For the purposes of this class, a `big' data set is one that is large enough that one cannot load the entire data set into a laptop's memory at once.
+Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
+My goal is to choose data sets that are comfortably over this threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes many hours.
+Practically speaking, from 2019 until the present I try to choose data sets that are on the order of 10 or 100 GB.
+
 \end{document}

### 2025-01-08 paper.tex
.

@@
 \usepackage[utf8]{inputenc}
+
+% for \url{}
+\usepackage{hyperref}
 
@@
 
-I’ve taught an upper division statistics ``Big Data'' course at UC Davis and CSU Sacramento and have gained some perspective on what works well for this course.
+I've taught an upper division statistics ``Big Data'' course at UC Davis and CSU Sacramento and have gained some perspective on what works well for this course.
 My goals with the course are to get students comfortable with using remote machines and to do realistic analysis of data sets that don’t fit in memory, typically on the order of 100GB or so.
@@
 
+I developed and have taught the course Stat 129, titled `Analyzing and Processing Big Data'.
+The catalog course description is:
+
+\emph{
+Statistical analysis of large, complex data sets.
+Topics include memory efficient data processing, the split-apply-combine strategy, rewriting programs for scalability, handling complex data formats, and applications such as statistical learning, dimension reduction, and efficient data representation.
+Students will access data and run code on remote servers.
+}
+
+Many of the techniques that we teach in this class work equally well on a modern laptop as on a server.
+
+It's essential that students access data and run code on remote servers, because there is a qualitative difference between running code on a laptop that's sitting in front of you and a machine that you never actually see, that you can only access through ssh, secure shell.
+I teach bash, the classic command line, because it has proven to be a durable technology; it's been around since the 1970's, and it isn't going away any time soon.
+In general, students benefit when instructors focus on proven core technologies, rather than on what happens to be popular this year, because they come away with skills that they can apply in many different contexts.
+
+My goal with the command line is for them to do it enough that they get over the initial fear that comes from moving from a Graphical User Interface (GUI), so they will be more comfortable in the future.
+Thus we need an actual remote server that students can log in to so that they can perform the tasks.
+The problem is that my job doesn't provide such a server.
+If I want students to have this experience, then it is on me to find and manage such a server.
+You may be in a similar situation of wanting to teach a class that relies on a server, and are wondering how should you do it?
+If so, read on.
+
+\Section{Which Data?}
+
 The class is called `Big Data'.
@@
 Practically speaking, from 2019 until the present I try to choose data sets that are on the order of 10 or 100 GB.
+The data analysis tasks will typically take 1 to 10 minutes if done in serial, and will be an order of magnitude faster if done in parallel.
+
+While teaching this course, I have to select data sets.
+Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
+In general, I'm looking for the following characteristics:
+\begin{enumerate}
+\item \textbf{Publicly Available} data should be available for legal public download.
+\item \textbf{Size} large enough to motivate the computational techniques we're learning, but no larger.
+\item \textbf{Format} common format, at least one tabular and another nested. Typically we will do a CSV/TSV for the tabular, and XML or JSON for the nested.
+\item \textbf{Background Knowledge} a college student should be able to understand what the data actually represents with minimal effort.
+\end{enumerate}
+
+Most of these principles I've violated at one time or another, and the difficulties became clear later.
+
+Amazon Open Data \url{https://registry.opendata.aws/} has provided a reliably good source of large data sets fitting these criterion. 
+Another good source is the US government \url{https://data.gov/}.
+
+Once I find a data set to work with, I download it to the server and massage it into the form where I want my students to see.
+A large part of the course is `cleaning' the data, preparing it for analysis by filtering, imputing, etc.
+TODO: cite
+However, it requires judgement as to which cleaning tasks are good for the students learning, and which are merely tedious or too uncommon.
+
+\subsection{server options}
+
+I'm an Assistant Professor in the Mathematics and Statistics Department at CSU Sacramento, a regional university.
+I have some professional background with computers and programming.
+In particular, I've done some server administration, and that experience has definitely proven useful when teaching this class.
+
+Our department plans to offer Stat 129 once a year for the foreseeable future, so we need a long term solution for a server.
+In a particular year we'll have one instructor, 15-25 students, and no TA's or other support.
+
+I don't need a large machine for any purpose other than teaching.
+
+What follows are the server options, ordered by my personal preference.
+
+\section{Best option - Dedicated support}
+
+If you're in the lucky position of having compute infrastructure with dedicated support, then I encourage you to exhaust
 

### 2025-01-08 paper.tex
.

@@
 In a particular year we'll have one instructor, 15-25 students, and no TA's or other support.
-
 I don't need a large machine for any purpose other than teaching.
+Please keep in mind the context I'm coming from when making the following recommendations, as they may not be appropriate for your situation.
 
@@
 
-\section{Best option - Dedicated support}
+What do you actually need to teach a class like this?
 
-If you're in the lucky position of having compute infrastructure with dedicated support, then I encourage you to exhaust
+\section{Best option - Dedicated Support and Infrastructure}
+
+If you're in the lucky position of having compute infrastructure with dedicated support in your department, college, or university, then this is an ideal option for teaching classes.
+You simply ask the system administrator for what you need, and they'll make it happen.
+The only hard part is knowing what it is you need.
+
+While teaching at UC Davis, the system administrator's helped me with the following tasks related to the class:
+\begin{itemize}
+\item Creating user accounts from a student roster.
+\item Installing R, Python, along with various packages and libraries of software.
+\item Setting up and configuring an open source database (Postgres) for use by students.
+\end{itemize}
+
+A systems administrator can do many other things, of course.
+Just ask!
+
+I feel that timely support is much more important the physical compute power, especially if the class is small.
+
+\section{Option 1 - Personal Physical Server}
+
+A personal physical server is either in a rack in a server room somewhere on campus, or else a large box in an office.
+If it's in a server rack then you can generally get a machine with higer specs, better internet connectivity, and it's not cluttering up your space.
+If the machine is in your office, then when you launch a big job the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which can be fun.
+
+This is my current solution, after having explored and experienced the others described in this article.
+I find that it is the most pragmatic solution, requiring the least effort on my part.
+
+The major advantages of this approach are cost and simplicity.
+Regarding cost, I've found it easier to fund a one time purchase of a few thousand dollars compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
+It's simple because you can set it up one time, and then use the same machine for several years.
+I myself am the systems administrator, so I can install whatever I want, whenever I want.
+I didn't need any specialized knowledge beyond the basics.
+I don't have to worry about any creating and maintaining cloud images.
+The downside of this is that I am my own support.
+If I spill coffee on your machine, then I'm in trouble.
+If something goes wrong, I'm the only one who's going to fix it.
+
+For a class of 15 students, I found that a single 32 core CPU with 128 GB of memory was just sufficient for students to experience the class assignments in the way I intended.
+I always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they are expecting.
+In practice it happened occassionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because I had warned them many times and they knew to expect this.
+With more students, I believe that students would feel crowded on the server.
+
+One technical detail to be aware of is that you must provide remote access so that others can log on to the machine.
+Campus IT handled this for me.
+I simply asked them for a hostname, and they took care of it.
+Our students need to be either on campus or at home on the campus VPN (Virtual Private Network), and they can login via ssh, for example: \texttt{ssh fitzgerald@nsm-stats.csus.ed}.
+I would recommend against setting up a physical server at your home, because this step could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
+
+The physical server feels out of date, and lacks the cachet of being `the cloud'.
+Despite that, it provides a legitimate and simple way to get students using a remote server.
+From a student's perspective, there's no difference between \texttt{ssh} to this machine, or to a virtual machine that you own that's running on the cloud.
 

### 2025-01-09 paper.tex
.

@@
-\documentclass{article}
+\documentclass[12pt]{article}
+
+\usepackage{geometry}
+\geometry{margin=1in}
+
 \usepackage[utf8]{inputenc}
@@
 
-\Section{Which Data?}
+\section{Which Data?}
 
@@
 In practice it happened occassionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because I had warned them many times and they knew to expect this.
+I've used a 16 core CPU in previous classes, and found that it was not sufficient.
 With more students, I believe that students would feel crowded on the server.
@@
 
+With either cloud option, you have two ways to provide students with access to a server: they can create their own virtual machines, or they can use a machine that you provide.
+If the students can have their own account in the cloud, then they can create their own virtual machines which they have complete control over, and can delete when they are done.
+They are the administrators to these machines with \texttt{sudo} access, so they can do anything they wish, and any mistakes they make won't affect anyone else.
+This would be a good option if the course is designed to teach them more about infrastructure, for example, you could allow them to implement their own web applications.
+
+With the second option, you create a virtual machine that's sufficiently large to accommadate all of the students simultaneously.
+You can either give the students SSH access, or allow them to access through a web interface, for example with JupyterHub or RStudio Cloud.
+The downside here is that you need to pay for a large enough machine all the time, or else manually load balance by switching to smaller or larger machines as the demand changes.
+
+\section{Option 2 - Public Cloud}
+
+I have had generally an excellent experience using the publicly funded cloud, Jetstream2.
+The two big benefits of this model are that it's free for the students, and it's a relatively simple way for the students to get real cloud experience..
+In this model, you apply for an allocation, Jetstream reviews and awards you a certain number of compute hours, you set up the accounts.
+To do the work, each student creates their own virtual machine and uses it to do the data analysis.
+I applied, was approved, and had everything set up in less than a week.
+
+The website \url{https://jetstream-cloud.org/index.html} states:
+TODO: cite
+
+\begin{quote}
+Jetstream2 makes cutting-edge high-performance computing and software easy to use for research regardless of your project’s scale, even if you have limited experience with supercomputing systems.
+
+Cloud-based and on-demand, the Jetstream2 environment also includes discipline-specific apps.
+You can even create virtual machines that look and feel like your lab workstation or home machine—but with thousands of times the computing power.
+\end{quote}
+
+The Jetstream2 user interface and options are far simpler than commercial cloud providers.
+
+Two disadvantages of Jetstream2 for teaching are the security model, and the lack of safeguards.
+Jetstream2 seems to be focused on allowing individual researchers and small groups, and enabling them to do projects with more compute than they would otherwise be able to.
+What I did, teaching a class where every student gets access, is certainly not the primary use case.
+When you're awarded an allocation, every user gets complete access to every part of the account.
+Students have administrative access to everything the instructor creates, and every other students machine.
+There are no limits on how much compute an individual can use, so it's possible for one user to use up all the credits.
+In other words, the only thing that's stopping them from doing something bad is their own awareness and respect for others.
+I mostly got around these issues by setting very clear expectations for behavior, with the understanding that this whole idea is based on trust.
+Even so, a student unintentionally left a large machine running for several weeks.
+I wasn't dilligently monitoring, so this used up around 25\% of our compute credits.
+Fortunately, we still had plenty to get through the course.
+
+A technical detail you'll have to handle is how students will access the large data sets.
+Jetstream2 provides data storage through the Manila Filesystems as a service, and this worked well for sharing large data sets across several different running instances.
+
+Another disadvantage for my use case is the uncertainty of how many years I'll be able to rely on it to continue offering the course, as I need to renew it annually.
+Jetstream2 is not designed as a long term solution for institutions offering the same course year after year. 
+I expect they would renew every year, but there's no guarantee of this happening.
+
+\section{Option 3 - Public Cloud}
+
+I've mostly
+
 \end{document}

### 2025-01-14 paper.tex
.

@@
 
+\section{Introduction}
+
+% TODO: Context, definitions, goals of the course, benefits of such a course to students
+
 I developed and have taught the course Stat 129, titled `Analyzing and Processing Big Data'.
@@
 }
+
+Our general approach to data science education is to provide students with real world data sets and ask open ended questions, squarely in the tradition of Deborah Nolan and Duncan Temple Lang \cite{nolan2010computing}.
+Indeed, they were our teachers and mentors.
 
@@
 
-\section{Which Data?}
+\subsection{Which Data?}
 
@@
 
-\subsection{server options}
+\section{server options}
 
@@
 
-\section{Best option - Dedicated Support and Infrastructure}
+\subsection{Best option - Dedicated Support and Infrastructure}
 
@@
 
+I've used more traditional supercomputers with job queueing software (SLURM) for teaching this class, but I feel that this is not ideal for the class, because of the amount of overhead, and the abstraction.
+For example, one topic that we cover is reading the output of \texttt{top/htop} to identify a bottleneck in a shell pipeline that's streaming hundreds of GB of data.
+It's possible to do this with job queueing software where the job is running on a worker node, but it takes either more advanced knowledge, or something specific to that system.
+Whenever possible I try to teach the tools that are the most common and widely applicable, to better serve the students in their future careers.
+
 I feel that timely support is much more important the physical compute power, especially if the class is small.
 
-\section{Option 1 - Personal Physical Server}
+\subsection{Option 1 - Personal Physical Server}
 
@@
 With either cloud option, you have two ways to provide students with access to a server: they can create their own virtual machines, or they can use a machine that you provide.
-If the students can have their own account in the cloud, then they can create their own virtual machines which they have complete control over, and can delete when they are done.
+I've used both of these options, and they both have a place, depending on the goals of the class.
+
+The first option provides more true `cloud' experience.
+The students can have their own account with the cloud provider, they create their own virtual machines which they have complete control over, and they delete them when they are done.
 They are the administrators to these machines with \texttt{sudo} access, so they can do anything they wish, and any mistakes they make won't affect anyone else.
@@
 
-With the second option, you create a virtual machine that's sufficiently large to accommadate all of the students simultaneously.
+The second option allows students to focus more on the data analysis rather than the infrastructure.
+You, the instructor, create a virtual machine that's sufficiently large to accommadate all of the students simultaneously.
 You can either give the students SSH access, or allow them to access through a web interface, for example with JupyterHub or RStudio Cloud.
@@
 
-\section{Option 2 - Public Cloud}
+\subsection{Option 2 - Public Cloud}
 
@@
 I expect they would renew every year, but there's no guarantee of this happening.
+If they choose not to renew, then I'll need to look for another solution.
 
-\section{Option 3 - Public Cloud}
+\subsection{Option 3 - Commercial Cloud}
 
-I've mostly
+Several companies offer commercial cloud services.
+I focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, so that our students can get marketable experience that they can then put on their resume.
+My favorite feature of AWS is the direct access to Amazon's high quality open data sets through the \texttt{aws s3} command line tools.
+From an instructor's perspective, this eliminates the step of downloading the data and making it available to students.
+
+The financial aspect is not ideal for students.
+These are commercial services, so a student must have a credit card and be prepared to pay for any compute time and storage that they use.
+In Spring 2022, I found that students on AWS can generally get the data processing experience for \$20 or less in one semester.
+If they make a mistake, say using an expensive service and forgetting to shut it off, then they can easily run up an expensive and unexpected bill that can be tough for a college student to cover.
+In the past, AWS offered an educational program with free credits for students.
+The big selling point for me was that students didn't need to even provide a credit card.
+I used this program, but they discontinued it, leaving me searching for other options.
+For a busy faculty member, this is a waste.
+
+Complexity and vendor specificity are also problems when teaching with cloud services.
+Amazon's EC2 (cloud compute server) ``provides a wide selection of over 750 instance types optimized to fit different use cases.''
+There are also hundreds of different services and several world regions to choose from.
+The services often have company specific names, which makes it difficult for an instructor to understand what they are.
+Indeed, the main purpose of the `AWS Cloud Practitioner' qualification seems to be to understand all of these proprietary names and services.
+This vendor specific education is all well and good if Amazon is providing the training, however, at a public institution my goal is to focus on the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
+
+While I feel that the commercial cloud services aren't currently the best choice for me, they might be the best choice for you.
+If you already have experience with a particular cloud provider, or you wish to build that experience, then go for it.
+If your institution has some connection with a cloud provider, then it may be mutually beneficial to leverage those connections.
+For example, the company may provide cloud services for free in exchange for the ability to recruit graduates into that company.
+
+I do believe that viable options can be built using commercial cloud services.
+For example, RStudio, Anaconda, and Snowflake offer cloud services that are essentially a higher level abstraction and user interface catered towards the data science R, Python, or SQL user.
+
+\section{Future Steps}
+
+If it was easier to set up and run a server, then I believe that more instructors would be interested in teaching this kind of class.
+For a lone instructor with no support, the barriers to entry are significant.
+I would welcome a solution that provides this capability publicly.
+It must have the following features:
+\begin{enumerate}
+\item \textbf{command line access} If this is present, then it should be possible to install and use all manner of command line software, including R, Python, and SQL.
+\item \textbf{100 GB data storage} 
+\item \textbf{free or low cost} Free for students is ideal, of course.
+    If that's not possible, then the cost over the semester should not exceed that of a reasonably priced textbook.
+\end{enumerate}
+Based on our observations above, there are several other features that would be nice to have, and I list them here in order of preference.
+
+\begin{enumerate}
+\item \textbf{longevity} It's much easier for instructors to invest in a particular solution if they can be sure that it will be around in 5 years.
+\end{enumerate}
+
+\bibliographystyle{plain}
+\bibliography{references}
 

### 2025-01-15 paper.tex
.

@@
 
-I've taught an upper division statistics ``Big Data'' course at UC Davis and CSU Sacramento and have gained some perspective on what works well for this course.
-My goals with the course are to get students comfortable with using remote machines and to do realistic analysis of data sets that don’t fit in memory, typically on the order of 100GB or so.
-
-I’ve tried what feels like an excessive number of ways to run the server for the students to get the `server experience':
-Campus supported Linux cluster (SLURM)
-AWS EC2 - student accounts and Jupyter Notebooks
-Google colab
-NSF Jetstream cloud (Access)
-Physical server in the corner of my office (current solution!)
+    It's a useful skill for students in data science to analyze and process large data sets located on remote servers.
+    To teach students these skills, we need to provide them with an authentic experience of logging in to a remote server and using it to answer questions based on a real world, large data set.
+    We offer practical suggestions to the instructor interested in using servers to teach data analysis, and share our experience using a variety of server options to teach this content, both with and without dedicated institutional support.
+    We compare and contrast different options including public and private cloud services, and personal physical machines.
+    Finally, we suggest a more sustainable sharing model that will allow more instructors to teach such classes without the overhead of having to manage their own server.
 
@@
 
-I developed and have taught the course Stat 129, titled `Analyzing and Processing Big Data'.
+\subsection{The Course}
+
+We developed and have taught the course Stat 129, titled `Analyzing and Processing Big Data'.
+This is an upper division elective course offered by the Mathematics and Statistics Department, and is required for students pursuing a bachelor's degree in Statistics with an emphasis in Data Science.
 The catalog course description is:
@@
 
+Students typically take this course their junior year after taking Stat 128, Statistical Computing, an upper division requirement for the statistics major focused on the R programming language, with content similar to Data Science in a Box, R4DS TODO: Cite.
+
+In Stat 129 we use and teach technologies including bash, Python, and SQL.
+Our intention is to expose students broadly to the common tools in data science.
 Our general approach to data science education is to provide students with real world data sets and ask open ended questions, squarely in the tradition of Deborah Nolan and Duncan Temple Lang \cite{nolan2010computing}.
@@
 
-Many of the techniques that we teach in this class work equally well on a modern laptop as on a server.
+\subsection{Why a Server?}
 
 It's essential that students access data and run code on remote servers, because there is a qualitative difference between running code on a laptop that's sitting in front of you and a machine that you never actually see, that you can only access through ssh, secure shell.
-I teach bash, the classic command line, because it has proven to be a durable technology; it's been around since the 1970's, and it isn't going away any time soon.
+We teach bash, the classic command line, because it has proven to be a durable technology; it's been around since the 1970's, and it isn't going away any time soon.
 In general, students benefit when instructors focus on proven core technologies, rather than on what happens to be popular this year, because they come away with skills that they can apply in many different contexts.
 
-My goal with the command line is for them to do it enough that they get over the initial fear that comes from moving from a Graphical User Interface (GUI), so they will be more comfortable in the future.
+Our goal with the command line is for them to do it enough that they get over the initial fear that comes from moving from a Graphical User Interface (GUI), so they will be more comfortable in the future.
 Thus we need an actual remote server that students can log in to so that they can perform the tasks.
-The problem is that my job doesn't provide such a server.
-If I want students to have this experience, then it is on me to find and manage such a server.
+The problem is that our job doesn't provide such a server.
+If we want students to have this experience, then it is on me to find and manage such a server.
 You may be in a similar situation of wanting to teach a class that relies on a server, and are wondering how should you do it?
 If so, read on.
+
+Incidentally, many of the data processing techniques that we teach in this class work equally well on a laptop as on a server.
 
@@
 Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
-My goal is to choose data sets that are comfortably over this threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes many hours.
-Practically speaking, from 2019 until the present I try to choose data sets that are on the order of 10 or 100 GB.
-The data analysis tasks will typically take 1 to 10 minutes if done in serial, and will be an order of magnitude faster if done in parallel.
+Our goal is to choose data sets that are comfortably over this threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes many hours.
+Practically speaking, a typical student laptop will have between 4 and 32 GB of memory at the time of this writing, and data sets that are on the order of 10-100 GB on disk work well to motivate the techniques.
+We assign common data analysis tasks that typically take 1 to 10 minutes if done in serial, and will be an order of magnitude faster if done in parallel.
 
-While teaching this course, I have to select data sets.
+While teaching this course, we have to select data sets.
 Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
@@
 
-Most of these principles I've violated at one time or another, and the difficulties became clear later.
+Most of these principles We've violated at one time or another, and the difficulties became clear later.
 
@@
 
-Once I find a data set to work with, I download it to the server and massage it into the form where I want my students to see.
+Once we find a data set to work with, we download it to the server and massage it into the form where we want our students to see.
 A large part of the course is `cleaning' the data, preparing it for analysis by filtering, imputing, etc.
@@
 
-\section{server options}
+\section{Server Options}
 
+% Introduce
+
+Which server option is best depends on the goals of the course, the level of institutional support, the number of students enrolled, the background of the students, and the background of the instructor.
+
+Here's our background.
 I'm an Assistant Professor in the Mathematics and Statistics Department at CSU Sacramento, a regional university.
-I have some professional background with computers and programming.
-In particular, I've done some server administration, and that experience has definitely proven useful when teaching this class.
+We have some professional background with computers and programming.
+In particular, We've done some server administration, and that experience has definitely proven useful when teaching this class.
 
@@
 In a particular year we'll have one instructor, 15-25 students, and no TA's or other support.
-I don't need a large machine for any purpose other than teaching.
+We don't need a large machine for any purpose other than teaching.
 Please keep in mind the context I'm coming from when making the following recommendations, as they may not be appropriate for your situation.
 
-What follows are the server options, ordered by my personal preference.
+What follows are the server options, ordered by our personal preference.
 
@@
 
-I've used more traditional supercomputers with job queueing software (SLURM) for teaching this class, but I feel that this is not ideal for the class, because of the amount of overhead, and the abstraction.
+We've used more traditional supercomputers with job queueing software (SLURM) for teaching this class, but we feel that this is not ideal for the class, because of the amount of overhead, and the abstraction.
 For example, one topic that we cover is reading the output of \texttt{top/htop} to identify a bottleneck in a shell pipeline that's streaming hundreds of GB of data.
 It's possible to do this with job queueing software where the job is running on a worker node, but it takes either more advanced knowledge, or something specific to that system.
-Whenever possible I try to teach the tools that are the most common and widely applicable, to better serve the students in their future careers.
+Whenever possible we try to teach the tools that are the most common and widely applicable, to better serve the students in their future careers.
 
-I feel that timely support is much more important the physical compute power, especially if the class is small.
+We feel that timely support is much more important the physical compute power, especially if the class is small.
 
@@
 
-This is my current solution, after having explored and experienced the others described in this article.
-I find that it is the most pragmatic solution, requiring the least effort on my part.
+This is our current solution, after having explored and experienced the others described in this article.
+We find that it is the most pragmatic solution, requiring the least effort on our part.
 
 The major advantages of this approach are cost and simplicity.
-Regarding cost, I've found it easier to fund a one time purchase of a few thousand dollars compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
+Regarding cost, We've found it easier to fund a one time purchase of a few thousand dollars compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
 It's simple because you can set it up one time, and then use the same machine for several years.
-I myself am the systems administrator, so I can install whatever I want, whenever I want.
-I didn't need any specialized knowledge beyond the basics.
-I don't have to worry about any creating and maintaining cloud images.
-The downside of this is that I am my own support.
-If I spill coffee on your machine, then I'm in trouble.
+We myself am the systems administrator, so we can install whatever we want, whenever we want.
+We didn't need any specialized knowledge beyond the basics.
+We don't have to worry about any creating and maintaining cloud images.
+The downside of this is that we am our own support.
+If we spill coffee on your machine, then I'm in trouble.
 If something goes wrong, I'm the only one who's going to fix it.
 
-For a class of 15 students, I found that a single 32 core CPU with 128 GB of memory was just sufficient for students to experience the class assignments in the way I intended.
-I always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they are expecting.
-In practice it happened occassionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because I had warned them many times and they knew to expect this.
-I've used a 16 core CPU in previous classes, and found that it was not sufficient.
-With more students, I believe that students would feel crowded on the server.
+For a class of 15 students, we found that a single 32 core CPU with 128 GB of memory was just sufficient for students to experience the class assignments in the way we intended.
+We always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they are expecting.
+In practice it happened occassionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because we had warned them many times and they knew to expect this.
+We've used a 16 core CPU in previous classes, and found that it was not sufficient.
+With more students, we believe that students would feel crowded on the server.
 
@@
 Campus IT handled this for me.
-I simply asked them for a hostname, and they took care of it.
+We simply asked them for a hostname, and they took care of it.
 Our students need to be either on campus or at home on the campus VPN (Virtual Private Network), and they can login via ssh, for example: \texttt{ssh fitzgerald@nsm-stats.csus.ed}.
-I would recommend against setting up a physical server at your home, because this step could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
+We would recommend against setting up a physical server at your home, because this step could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
 
@@
 With either cloud option, you have two ways to provide students with access to a server: they can create their own virtual machines, or they can use a machine that you provide.
-I've used both of these options, and they both have a place, depending on the goals of the class.
+We've used both of these options, and they both have a place, depending on the goals of the class.
 
@@
 
-I have had generally an excellent experience using the publicly funded cloud, Jetstream2.
+We have had generally an excellent experience using the publicly funded cloud, Jetstream2.
 The two big benefits of this model are that it's free for the students, and it's a relatively simple way for the students to get real cloud experience..
@@
 To do the work, each student creates their own virtual machine and uses it to do the data analysis.
-I applied, was approved, and had everything set up in less than a week.
+We applied, was approved, and had everything set up in less than a week.
 
@@
 Jetstream2 seems to be focused on allowing individual researchers and small groups, and enabling them to do projects with more compute than they would otherwise be able to.
-What I did, teaching a class where every student gets access, is certainly not the primary use case.
+What we did, teaching a class where every student gets access, is certainly not the primary use case.
 When you're awarded an allocation, every user gets complete access to every part of the account.
@@
 In other words, the only thing that's stopping them from doing something bad is their own awareness and respect for others.
-I mostly got around these issues by setting very clear expectations for behavior, with the understanding that this whole idea is based on trust.
+We mostly got around these issues by setting very clear expectations for behavior, with the understanding that this whole idea is based on trust.
 Even so, a student unintentionally left a large machine running for several weeks.
-I wasn't dilligently monitoring, so this used up around 25\% of our compute credits.
+We wasn't dilligently monitoring, so this used up around 25\% of our compute credits.
 Fortunately, we still had plenty to get through the course.
@@
 
-Another disadvantage for my use case is the uncertainty of how many years I'll be able to rely on it to continue offering the course, as I need to renew it annually.
+Another disadvantage for our use case is the uncertainty of how many years I'll be able to rely on it to continue offering the course, as we need to renew it annually.
 Jetstream2 is not designed as a long term solution for institutions offering the same course year after year. 
-I expect they would renew every year, but there's no guarantee of this happening.
+We expect they would renew every year, but there's no guarantee of this happening.
 If they choose not to renew, then I'll need to look for another solution.
@@
 Several companies offer commercial cloud services.
-I focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, so that our students can get marketable experience that they can then put on their resume.
-My favorite feature of AWS is the direct access to Amazon's high quality open data sets through the \texttt{aws s3} command line tools.
+We focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, so that our students can get marketable experience that they can then put on their resume.
+Our favorite feature of AWS is the direct access to Amazon's high quality open data sets through the \texttt{aws s3} command line tools.
 From an instructor's perspective, this eliminates the step of downloading the data and making it available to students.
@@
 These are commercial services, so a student must have a credit card and be prepared to pay for any compute time and storage that they use.
-In Spring 2022, I found that students on AWS can generally get the data processing experience for \$20 or less in one semester.
+In Spring 2022, we found that students on AWS can generally get the data processing experience for \$20 or less in one semester.
 If they make a mistake, say using an expensive service and forgetting to shut it off, then they can easily run up an expensive and unexpected bill that can be tough for a college student to cover.
@@
 The big selling point for me was that students didn't need to even provide a credit card.
-I used this program, but they discontinued it, leaving me searching for other options.
+We used this program, but they discontinued it, leaving me searching for other options.
 For a busy faculty member, this is a waste.
@@
 Indeed, the main purpose of the `AWS Cloud Practitioner' qualification seems to be to understand all of these proprietary names and services.
-This vendor specific education is all well and good if Amazon is providing the training, however, at a public institution my goal is to focus on the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
+This vendor specific education is all well and good if Amazon is providing the training, however, at a public institution our goal is to focus on the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
 
-While I feel that the commercial cloud services aren't currently the best choice for me, they might be the best choice for you.
+While we feel that the commercial cloud services aren't currently the best choice for me, they might be the best choice for you.
 If you already have experience with a particular cloud provider, or you wish to build that experience, then go for it.
@@
 
-I do believe that viable options can be built using commercial cloud services.
+We do believe that viable options can be built using commercial cloud services.
 For example, RStudio, Anaconda, and Snowflake offer cloud services that are essentially a higher level abstraction and user interface catered towards the data science R, Python, or SQL user.
@@
 
-If it was easier to set up and run a server, then I believe that more instructors would be interested in teaching this kind of class.
+If it was easier to set up and run a server, then we believe that more instructors would be interested in teaching this kind of class.
 For a lone instructor with no support, the barriers to entry are significant.
-I would welcome a solution that provides this capability publicly.
+We would welcome a solution that provides this capability publicly.
 It must have the following features:
@@
 \end{enumerate}
-Based on our observations above, there are several other features that would be nice to have, and I list them here in order of preference.
+Based on our observations above, there are several other features that would be nice to have, and we list them here in order of preference.
 
@@
 
+\section{Scratch}
+
+We've taught an upper division statistics ``Big Data'' course at UC Davis and CSU Sacramento and have gained some perspective on what works well for this course.
+Our goals with the course are to get students comfortable with using remote machines and to do realistic analysis of data sets that don’t fit in memory, typically on the order of 100GB or so.
+
+I’ve tried what feels like an excessive number of ways to run the server for the students to get the `server experience':
+Campus supported Linux cluster (SLURM)
+AWS EC2 - student accounts and Jupyter Notebooks
+Google colab
+NSF Jetstream cloud (Access)
+Physical server in the corner of our office (current solution!)
+
 \end{document}

### 2025-01-28 paper.tex
revise abstract

@@
 \usepackage{geometry}
-\geometry{margin=1in}
+%\geometry{margin=1in}
+
+\usepackage{setspace}
+\doublespacing
 
@@
 
-    It's a useful skill for students in data science to analyze and process large data sets located on remote servers.
-    To teach students these skills, we need to provide them with an authentic experience of logging in to a remote server and using it to answer questions based on a real world, large data set.
-    We offer practical suggestions to the instructor interested in using servers to teach data analysis, and share our experience using a variety of server options to teach this content, both with and without dedicated institutional support.
-    We compare and contrast different options including public and private cloud services, and personal physical machines.
-    Finally, we suggest a more sustainable sharing model that will allow more instructors to teach such classes without the overhead of having to manage their own server.
+    We compare and contrast different server options for providing undergraduate students with an authentic experience analyzing large, real world data sets located on powerful remote servers.
+These server options include institutionally supported academic computers, commercial cloud services, public cloud services, and personal physical machines.
+% TODO: Align these options with the sections
+We briefly describe the course, and offer suggestions on when to choose which server option, based on the background of the instructor and the level of institutional support.
+Looking to the future, we suggest requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having to manage and administer their own server.
 

### 2025-01-28 paper.tex
section 1

@@
 
-\section{Introduction}
-
 % TODO: Context, definitions, goals of the course, benefits of such a course to students
 
-\subsection{The Course}
+\section{The Course}
 
-We developed and have taught the course Stat 129, titled `Analyzing and Processing Big Data'.
-This is an upper division elective course offered by the Mathematics and Statistics Department, and is required for students pursuing a bachelor's degree in Statistics with an emphasis in Data Science.
+Stat 129, `Analyzing and Processing Big Data', is an upper division elective course offered by the Mathematics and Statistics Department at CSU Sacramento, and is required for students pursuing a bachelor's degree in Statistics with an Emphasis in Data Science.
 The catalog course description is:
@@
 
-Students typically take this course their junior year after taking Stat 128, Statistical Computing, an upper division requirement for the statistics major focused on the R programming language, with content similar to Data Science in a Box, R4DS TODO: Cite.
+Stat 129 uses technologies including bash, Python, and SQL to analyze large data sets, with the intention of exposing students to the common tools in data science.
+Students tend to be highly motivated, because many of them hope to land jobs requiring these skills after graduation.
+Stat 129 requires a programming prerequisite, since it is too advanced for a first programming course.
+Statistics majors take Stat 129 after taking `Statistical Computing', an upper division course focused on data analysis using the R programming language, with content similar to Data Science in a Box, R4DS TODO: Cite.
+%Ideally, students would have taken two semesters of programming before this course.
 
-In Stat 129 we use and teach technologies including bash, Python, and SQL.
-Our intention is to expose students broadly to the common tools in data science.
-Our general approach to data science education is to provide students with real world data sets and ask open ended questions, squarely in the tradition of Deborah Nolan and Duncan Temple Lang \cite{nolan2010computing}.
-Indeed, they were our teachers and mentors.
+The philosophy behind this class is to provide students with real world data sets and ask open ended questions, following in the footsteps of our teachers and mentors, Deborah Nolan and Duncan Temple Lang \cite{nolan2010computing}.
+We have been teaching data science programming technologies in an academic setting since 2015, and we developed Stat 129 and have taught it each year since 2021.
+Feedback from this data forward approach in general, and this class in particular, has been overwhelmingly positive.
+In Stat 129 students ask real questions based on real data that resides on a real server, and for this reason several alumni have reported that this has been the single most practical and valuable class that they have taken in college.
 
@@
 
-It's essential that students access data and run code on remote servers, because there is a qualitative difference between running code on a laptop that's sitting in front of you and a machine that you never actually see, that you can only access through ssh, secure shell.
+It's essential that students access data and run code on remote servers for a class on `Big Data', because there is a qualitative difference between running code on a laptop that's sitting in front of you and a machine that you never actually see, that you can only access through ssh, secure shell.
 We teach bash, the classic command line, because it has proven to be a durable technology; it's been around since the 1970's, and it isn't going away any time soon.

### 2025-01-28 paper.tex
why server section

@@
 
-It's essential that students access data and run code on remote servers for a class on `Big Data', because there is a qualitative difference between running code on a laptop that's sitting in front of you and a machine that you never actually see, that you can only access through ssh, secure shell.
-We teach bash, the classic command line, because it has proven to be a durable technology; it's been around since the 1970's, and it isn't going away any time soon.
-In general, students benefit when instructors focus on proven core technologies, rather than on what happens to be popular this year, because they come away with skills that they can apply in many different contexts.
+Students must access data and run code on remote servers for a class on `Big Data', because that's where big data is typically found: on remote servers.
+There is a qualitative difference between the experience of running code on a laptop with a monitor that's sitting in front of you, and on a machine that you never actually see, that you can only access through \texttt{ssh}, secure shell.
+Our goal with embracing the shell is to move students beyond the initial unfamiliarity and discomfort that users experience when they first move away from a Graphical User Interface (GUI).
+%Incidentally, many of the techniques for processing data larger than memory that we teach in this class work equally well on a modest laptop as on a powerful server.
+To access the large data sets and complete the assignments, quizzes, and exams, the students must login to the class server through \texttt{ssh}.
 
-Our goal with the command line is for them to do it enough that they get over the initial fear that comes from moving from a Graphical User Interface (GUI), so they will be more comfortable in the future.
-Thus we need an actual remote server that students can log in to so that they can perform the tasks.
-The problem is that our job doesn't provide such a server.
-If we want students to have this experience, then it is on me to find and manage such a server.
-You may be in a similar situation of wanting to teach a class that relies on a server, and are wondering how should you do it?
-If so, read on.
+We teach the classic shell bash, because it has proven to be a durable technology.
+If one hopes to do anything with more advanced tools such as larger compute clusters or frameworks such Hadoop, then a foundational knowledge of bash is essential.
+TODO: cite
+In general, students in technology oriented class such as Stat 129 benefit when instructors focus on ubiquitous and proven core technologies, rather than on what happens to be popular this year, because then they come away with skills that they can apply in many different contexts.
 
-Incidentally, many of the data processing techniques that we teach in this class work equally well on a laptop as on a server.
+%Thus we need an actual remote server that students can log in to so that they can perform the tasks.
 
@@
 Which server option is best depends on the goals of the course, the level of institutional support, the number of students enrolled, the background of the students, and the background of the instructor.
+
+%The problem is that our job doesn't provide such a server.
+%If we want students to have this experience, then it is on me to find and manage such a server.
+%You may be in a similar situation of wanting to teach a class that relies on a server, and are wondering how should you do it?
+%If so, read on.
 

### 2025-01-28 paper.tex
data section

@@
 
-The class is called `Big Data'.
-For this term to have any meaning, we need to define what we mean by `big'.
-For the purposes of this class, a `big' data set is one that is large enough that one cannot load the entire data set into a laptop's memory at once.
-Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
-Our goal is to choose data sets that are comfortably over this threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes many hours.
-Practically speaking, a typical student laptop will have between 4 and 32 GB of memory at the time of this writing, and data sets that are on the order of 10-100 GB on disk work well to motivate the techniques.
-We assign common data analysis tasks that typically take 1 to 10 minutes if done in serial, and will be an order of magnitude faster if done in parallel.
+%The term `Big Data' appears in the title for Stat 129; for this term to be more than a buzz word, we need to define it.
+% TODO: Could talk about how others use this term.
+For the purposes of Stat 129, `big data' means a data set that is large enough such that one cannot load the entire data set into a laptop's memory at once.
+In a typical introductory data analysis class, a student reads a single data file into R or Python and proceeds with the analysis steps \cite{dsbox}.
+This works great if the data fits into memory, but if the data does not fit into memory, then it simply doesn't work.
+Data sets that are larger than memory require entirely different approaches, such as streaming through to filter out a subset of interest.
+%Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
+Our goal is to choose data sets that are comfortably over this memory threshold, such that the techniques we're learning are motivated, but not gratuitously large such that answering a basic question takes an excessive amount of time.
+Practically speaking, at the time of this writing, a typical student laptop will have between 4 and 32 GB of memory, and data sets that are on the order of 10-100 GB on disk work well to motivate the techniques.
+We assign common data analysis tasks that typically take 1 to 10 minutes if done in serial, and are an order of magnitude faster if done in parallel.
 
-While teaching this course, we have to select data sets.
-Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
-In general, I'm looking for the following characteristics:
+%While teaching this course, we have to select data sets.
+%Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
+When teaching Stat 129, we look for data sets that satisfy the criteria below.
 \begin{enumerate}
-\item \textbf{Publicly Available} data should be available for legal public download.
-\item \textbf{Size} large enough to motivate the computational techniques we're learning, but no larger.
-\item \textbf{Format} common format, at least one tabular and another nested. Typically we will do a CSV/TSV for the tabular, and XML or JSON for the nested.
-\item \textbf{Background Knowledge} a college student should be able to understand what the data actually represents with minimal effort.
+\item \textbf{Publicly Available} The data should be available for legal public download.
+	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
+\item \textbf{Size} The files should be large enough to motivate the computational techniques we're learning, but no larger.
+\item \textbf{Format} The storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
+	Avoid proprietary, novel, or highly specialized data formats.
+	In one semester of Stat 129, we work with data in a plain text tabular format such as CSV or TSV, and nested formats such as XML or JSON.
+\item \textbf{Background Knowledge} A college student should be able to understand what the data actually represents.
+	If you cannot explain the essential structure and idea in a single class period, then that data is probably too complex.
 \end{enumerate}
-
-Most of these principles We've violated at one time or another, and the difficulties became clear later.
-
-Amazon Open Data \url{https://registry.opendata.aws/} has provided a reliably good source of large data sets fitting these criterion. 
+These criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on genomic data.
+%Most of these principles We've violated at one time or another, and the difficulties became clear later.
+Amazon Open Data \url{https://registry.opendata.aws/} has provided a reliably good source of large data sets fitting these criteria. 
 Another good source is the US government \url{https://data.gov/}.
 
-Once we find a data set to work with, we download it to the server and massage it into the form where we want our students to see.
-A large part of the course is `cleaning' the data, preparing it for analysis by filtering, imputing, etc.
-TODO: cite
-However, it requires judgement as to which cleaning tasks are good for the students learning, and which are merely tedious or too uncommon.
+We often find interesting data sets that are close, but not quite what we're looking for.
+In this case, it may be possible to process the data set before providing it to students to make it better suit the goals of the course.
+As a specific example, consider the tax return data on nonprofits provided by the Internal Revenue Service of the United States\cite{irs990}.
+We removed the XML namespace information in these files so that students learning XPATH can write \texttt{"//ActivityOrMissionDesc"} to find the `activity or mission description' for a particular nonprofit rather than the cumbersome \texttt{//*[local-name()='ActivityOrMissionDesc']}.
+The latter is distracting for students just learning XPATH.
+
+The large data sets are the heart of Stat 129.
+To properly analyze them we need to provide our students with a compute server.
 

### 2025-01-28 paper.tex
rearrange

@@
 
-\subsection{Why a Server?}
-
-Students must access data and run code on remote servers for a class on `Big Data', because that's where big data is typically found: on remote servers.
-There is a qualitative difference between the experience of running code on a laptop with a monitor that's sitting in front of you, and on a machine that you never actually see, that you can only access through \texttt{ssh}, secure shell.
-Our goal with embracing the shell is to move students beyond the initial unfamiliarity and discomfort that users experience when they first move away from a Graphical User Interface (GUI).
-%Incidentally, many of the techniques for processing data larger than memory that we teach in this class work equally well on a modest laptop as on a powerful server.
-To access the large data sets and complete the assignments, quizzes, and exams, the students must login to the class server through \texttt{ssh}.
-
-We teach the classic shell bash, because it has proven to be a durable technology.
-If one hopes to do anything with more advanced tools such as larger compute clusters or frameworks such Hadoop, then a foundational knowledge of bash is essential.
-TODO: cite
-In general, students in technology oriented class such as Stat 129 benefit when instructors focus on ubiquitous and proven core technologies, rather than on what happens to be popular this year, because then they come away with skills that they can apply in many different contexts.
-
-%Thus we need an actual remote server that students can log in to so that they can perform the tasks.
-
 \subsection{Which Data?}
@@
 
+\subsection{Why a Server?}
+
+Students must access data and run code on remote servers for a class on `Big Data', because that's where big data is typically found: on remote servers.
+There is a qualitative difference between the experience of running code on a laptop with a monitor that's sitting in front of you, and on a machine that you never actually see, that you can only access through \texttt{ssh}, secure shell.
+Our goal with embracing the shell is to move students beyond the initial unfamiliarity and discomfort that users experience when they first move away from a Graphical User Interface (GUI).
+%Incidentally, many of the techniques for processing data larger than memory that we teach in this class work equally well on a modest laptop as on a powerful server.
+To access the large data sets and complete the assignments, quizzes, and exams, the students must login to the class server through \texttt{ssh}.
+
+We teach the classic shell bash, because it has proven to be a durable technology.
+If one hopes to do anything with more advanced tools such as larger compute clusters or frameworks such Hadoop, then a foundational knowledge of bash is essential.
+TODO: cite
+In general, students in technology oriented class such as Stat 129 benefit when instructors focus on ubiquitous and proven core technologies, rather than on what happens to be popular this year, because then they come away with skills that they can apply in many different contexts.
+
+%Thus we need an actual remote server that students can log in to so that they can perform the tasks.
+
 \section{Server Options}
@@
 
-\subsection{Best option - Dedicated Support and Infrastructure}
+\subsection{Dedicated Support and Infrastructure}
 
@@
 
-\subsection{Option 1 - Personal Physical Server}
+\subsection{Personal Physical Server}
 
@@
 
-\subsection{Option 2 - Public Cloud}
+\subsection{Public Cloud}
 
@@
 
-\subsection{Option 3 - Commercial Cloud}
+\subsection{Commercial Cloud}
 
@@
 \end{enumerate}
-Based on our observations above, there are several other features that would be nice to have, and we list them here in order of preference.
+Based on our observations above, there are several other features that would be nice to have, and we list some of them here.
 

### 2025-01-28 paper.tex
coming along

@@
 \begin{enumerate}
-\item \textbf{Publicly Available} The data should be available for legal public download.
+\item \textbf{Size} The files should be large enough to motivate the computational techniques we're learning, but no larger, as described above.
+\item \textbf{Publicly Available} The data should be readily available for legal public download.
 	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
-\item \textbf{Size} The files should be large enough to motivate the computational techniques we're learning, but no larger.
 \item \textbf{Format} The storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
 	Avoid proprietary, novel, or highly specialized data formats.
-	In one semester of Stat 129, we work with data in a plain text tabular format such as CSV or TSV, and nested formats such as XML or JSON.
+    Each semester of Stat 129, we work with data in a plain text tabular format such as CSV or TSV, and nested formats such as XML or JSON.
 \item \textbf{Background Knowledge} A college student should be able to understand what the data actually represents.
-	If you cannot explain the essential structure and idea in a single class period, then that data is probably too complex.
+	If it takes longer than a single class period to explain the core idea of what the data captures, then that data is probably too complex.
 \end{enumerate}
@@
 %Most of these principles We've violated at one time or another, and the difficulties became clear later.
-Amazon Open Data \url{https://registry.opendata.aws/} has provided a reliably good source of large data sets fitting these criteria. 
-Another good source is the US government \url{https://data.gov/}.
+Amazon Open Data\footnote{\url{https://registry.opendata.aws/}} has provided a reliably good source of large data sets fitting these criteria. 
+The US government\footnote{\url{https://data.gov/}} is another good source.
 

### 2025-02-04 paper.tex
.

@@
 We teach the classic shell bash, because it has proven to be a durable technology.
-If one hopes to do anything with more advanced tools such as larger compute clusters or frameworks such Hadoop, then a foundational knowledge of bash is essential.
-TODO: cite
+Multiple alumni of Stat 129 have reported back that it was the command line skills that led directly to them getting a variety of jobs.
 In general, students in technology oriented class such as Stat 129 benefit when instructors focus on ubiquitous and proven core technologies, rather than on what happens to be popular this year, because then they come away with skills that they can apply in many different contexts.
+If one hopes to do anything with more advanced tools such as larger compute clusters or frameworks such Hadoop MapReduce \cite{white2012hadoop}, then a foundational knowledge of bash is essential.
 
-%Thus we need an actual remote server that students can log in to so that they can perform the tasks.
+A single server hits the sweet spot for Stat 129; clusters are less ideal, because of the overhead and the abstraction they introduce.
+We've used a cluster with SLURM \cite{yoo2003slurm} job queueing software, common in high performance computing (HPC), for teaching Stat 141C (Big Data and High Performance Computing) at UC Davis in 2019.
+One topic that we cover in Stat 129 is reading the output of \texttt{top/htop} to identify a bottleneck in a shell pipeline that's streaming hundreds of GB of data.
+It's possible to do this with job queueing software where the job is running on a worker node, but it takes either more advanced knowledge, or something specific to that system.
+Whenever possible we try to teach the tools that are the most common and widely applicable, to better serve the students in their future careers.
 
@@
 Which server option is best depends on the goals of the course, the level of institutional support, the number of students enrolled, the background of the students, and the background of the instructor.
+The overarching goal of Stat 129 is to do real things on real data sets, using professional software.
+Students use the server for every assignment and assessment for 15 weeks, so it makes sense to invest some time teaching them skills and tools with steeper learning curves, such as terminal based text editors.
+If a different class only used a server for two weeks of the semester to examine one particular data set, then it probably makes more sense to use a more approachable web browser based graphical user interface to the server such as Jupyter Notebooks \cite{kluyver2016jupyter}, which will inform the best choice of server options.
+
+Institutional support for teaching a class like Stat 129 varies widely.
+For us, there were no servers available to use for teaching at the department, college, or university level; however, the college did provide us with startup funds, which funded the experimentation that led to this paper.
+The Mathematics and Statistics Department at Sacramento State University plans to offer Stat 129 once a year for the foreseeable future, so we need a long term solution for a server.
+In a particular year we'll have one instructor, 15-30 students, and no staff, teaching assistants, or other support.
+We don't need a large machine for any purpose other than teaching.
+
+We have some professional background in programming, and our server administration experience has proven particularly useful for teaching Stat 129.
+What follows are the server options, ordered by our personal preference.
+There's no one size fits all solution, and what's best for you may well differ from our preferences.
 
@@
 
-Here's our background.
-I'm an Assistant Professor in the Mathematics and Statistics Department at CSU Sacramento, a regional university.
-We have some professional background with computers and programming.
-In particular, We've done some server administration, and that experience has definitely proven useful when teaching this class.
+%What do you actually need to teach a class like this?
 
-Our department plans to offer Stat 129 once a year for the foreseeable future, so we need a long term solution for a server.
-In a particular year we'll have one instructor, 15-25 students, and no TA's or other support.
-We don't need a large machine for any purpose other than teaching.
-Please keep in mind the context I'm coming from when making the following recommendations, as they may not be appropriate for your situation.
-
-What follows are the server options, ordered by our personal preference.
-
-What do you actually need to teach a class like this?
-
-\subsection{Dedicated Support and Infrastructure}
+\subsection{Dedicated Support}
 
 If you're in the lucky position of having compute infrastructure with dedicated support in your department, college, or university, then this is an ideal option for teaching classes.
-You simply ask the system administrator for what you need, and they'll make it happen.
-The only hard part is knowing what it is you need.
+You simply ask the system administrator for what you need, and they make it happen.
+The challenge for the instructor is determining what you need!
 
-While teaching at UC Davis, the system administrator's helped me with the following tasks related to the class:
+While teaching at UC Davis, the system administrator helped us with the following tasks related to the class:
 \begin{itemize}
+\item Selecting the right servers to offer to the class.
 \item Creating user accounts from a student roster.
@@
 \item Setting up and configuring an open source database (Postgres) for use by students.
+\item Troubleshooting connectivity and user account issues.
 \end{itemize}
-
-A systems administrator can do many other things, of course.
-Just ask!
-
-We've used more traditional supercomputers with job queueing software (SLURM) for teaching this class, but we feel that this is not ideal for the class, because of the amount of overhead, and the abstraction.
-For example, one topic that we cover is reading the output of \texttt{top/htop} to identify a bottleneck in a shell pipeline that's streaming hundreds of GB of data.
-It's possible to do this with job queueing software where the job is running on a worker node, but it takes either more advanced knowledge, or something specific to that system.
-Whenever possible we try to teach the tools that are the most common and widely applicable, to better serve the students in their future careers.
+Of these tasks, setting up the Postgres database took the most time and expertise, and we were grateful to have the support.
+% A system administrator can do many other things, of course.
+% Just ask!
 
@@
 We would recommend against setting up a physical server at your home, because this step could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
+
+It's relatively easy and direct to automate certain grading tasks on a physical server.
+For example...
 

### 2025-02-04 paper.tex
.

@@
 We don't need a large machine for any purpose other than teaching.
+We have some professional background in programming, and our server administration experience has proven particularly useful.
 
-We have some professional background in programming, and our server administration experience has proven particularly useful for teaching Stat 129.
 What follows are the server options, ordered by our personal preference.
@@
 You simply ask the system administrator for what you need, and they make it happen.
-The challenge for the instructor is determining what you need!
+The challenge for you as the instructor is determining what you need!
 
@@
 Of these tasks, setting up the Postgres database took the most time and expertise, and we were grateful to have the support.
-% A system administrator can do many other things, of course.
+A system administrator can do many other things, of course.
+For teaching, timely, helpful support is much more important than physical compute power.
 % Just ask!
-
-We feel that timely support is much more important the physical compute power, especially if the class is small.
 
@@
 
-A personal physical server is either in a rack in a server room somewhere on campus, or else a large box in an office.
-If it's in a server rack then you can generally get a machine with higer specs, better internet connectivity, and it's not cluttering up your space.
-If the machine is in your office, then when you launch a big job the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which can be fun.
+A personal physical server is essentially the same as a powerful desktop machine, except that you can access it remotely, and it doesn't need to have a monitor.
+The machine itself might be in a rack in a server room on campus, or in someone's office.
+If it's in a server rack then you can generally get a machine with more CPU's and memory, better internet connectivity, and it's not cluttering up your space.
+If the machine is in your office, then when you launch a big job, the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which is fun.
+A physical machine in our office is our current solution, after having explored and experienced the others described in this article.
+We find that it is the most pragmatic long term solution, requiring the least effort on our part.
 
-This is our current solution, after having explored and experienced the others described in this article.
-We find that it is the most pragmatic solution, requiring the least effort on our part.
+The major advantages of the personal physical server are cost and simplicity.
+Regarding cost, we've found it easier to fund a one time purchase of a few thousand dollars compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
+It's simple because you can set it up one time, and then use the same machine for several years.
+We didn't need any specialized knowledge beyond the basics.
+We completely own and administrator the machine, so we can install whatever we want, whenever we want.
+It's relatively easy and direct to automate certain grading tasks, for example, one can look at \texttt{history} to see every command a student typed during a quiz.
 
-The major advantages of this approach are cost and simplicity.
-Regarding cost, We've found it easier to fund a one time purchase of a few thousand dollars compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
-It's simple because you can set it up one time, and then use the same machine for several years.
-We myself am the systems administrator, so we can install whatever we want, whenever we want.
-We didn't need any specialized knowledge beyond the basics.
-We don't have to worry about any creating and maintaining cloud images.
-The downside of this is that we am our own support.
-If we spill coffee on your machine, then I'm in trouble.
-If something goes wrong, I'm the only one who's going to fix it.
+The disadvantage of a single physical machine is that if there is a problem, then it affects the entire class, and we are our own support.
+Students have made mistakes and filled up all the disk space, and we needed to diagnose and fix the issue on our own.
+Pedagogically, this mistake followed by the unusable server highlights the importance of not creating unnecessary copies when working with large amounts of data, and it makes students more aware of how much data they're saving.
+The fix is straightforward: just delete the big files that the students accidentally made copies of.
+A more proactive solution to this common problem is to impose a quota on disk usage.
+We're taking a risk that the server could fail in a more serious way, but it's a calculated risk.
 
-For a class of 15 students, we found that a single 32 core CPU with 128 GB of memory was just sufficient for students to experience the class assignments in the way we intended.
+%when the students make a mistake and fill up all the disk space, it does not affect anyone or anything outside of the class.
+%If we spill coffee on the machine, or it fails 
+%Nothing else essential is running on the machine, so 
+%If something goes wrong, I'm the only one who's going to fix it.
+%We don't have to worry about any creating and maintaining cloud images.
+
+The server doesn't need to be particularly powerful or expensive.
+For a class of 15 students, we found that a single 32 core CPU, a GPU, and 128 GB of memory was just sufficient for students to experience the class assignments in the way we intended.
+This machine cost around \$3,000 when purchased in 2022.
 We always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they are expecting.
-In practice it happened occassionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because we had warned them many times and they knew to expect this.
-We've used a 16 core CPU in previous classes, and found that it was not sufficient.
+In practice it happened occasionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because we had warned them many times and they knew to expect this.
+We've used a 16 core CPU machine in previous semesters, and found that it was not sufficient for 15 students.
 With more students, we believe that students would feel crowded on the server.
@@
 One technical detail to be aware of is that you must provide remote access so that others can log on to the machine.
-Campus IT handled this for me.
-We simply asked them for a hostname, and they took care of it.
-Our students need to be either on campus or at home on the campus VPN (Virtual Private Network), and they can login via ssh, for example: \texttt{ssh fitzgerald@nsm-stats.csus.ed}.
-We would recommend against setting up a physical server at your home, because this step could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
-
-It's relatively easy and direct to automate certain grading tasks on a physical server.
-For example...
+Campus IT handled this by providing a DNS hostname.
+Our students need to be either on campus or at home on the campus VPN (Virtual Private Network), and they can login via ssh, for example: \texttt{ssh fitzgerald@nsm-stats.csus.edu}.
+We recommend against setting up a physical server at your home, because providing reliable remote access could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
 
@@
 
-With either cloud option, you have two ways to provide students with access to a server: they can create their own virtual machines, or they can use a machine that you provide.
+\subsection{Cloud}
+
+In Stat 129, students process data through a virtual machine (VM), a logical rather than a physical computer.
+We've experimented with two ways of providing students with cloud access: administrative accounts on the cloud, or local user accounts on a VM.
 We've used both of these options, and they both have a place, depending on the goals of the class.
 
-The first option provides more true `cloud' experience.
-The students can have their own account with the cloud provider, they create their own virtual machines which they have complete control over, and they delete them when they are done.
+Administrative cloud accounts provide students with a richer cloud experience, as they create their own virtual machines which they have complete control over.
 They are the administrators to these machines with \texttt{sudo} access, so they can do anything they wish, and any mistakes they make won't affect anyone else.
-This would be a good option if the course is designed to teach them more about infrastructure, for example, you could allow them to implement their own web applications.
+Load balancing has the potential to work better in this scenario, as students can create machines that are as large as they like, and delete them when they are done.
+This would be a good option if the course is designed to teach them more about infrastructure, for example, how to implement a full stack web application.
+This rich experience does come with some additional complexity and responsibility, for example, students need to load and save their results as they create and delete machines.
 
-The second option allows students to focus more on the data analysis rather than the infrastructure.
-You, the instructor, create a virtual machine that's sufficiently large to accommadate all of the students simultaneously.
-You can either give the students SSH access, or allow them to access through a web interface, for example with JupyterHub or RStudio Cloud.
+Local user accounts on a single VM allow students to focus more on the data analysis rather than the cloud ideas and infrastructure.
+In this scenario the instructor creates a VM that's large enough to accommodate all of the students simultaneously, and students access through SSH or a web interface.
 The downside here is that you need to pay for a large enough machine all the time, or else manually load balance by switching to smaller or larger machines as the demand changes.
+If we were goint
 
-\subsection{Public Cloud}
+Note that providing students with administrative cloud accounts is not mutually exclusive from simpler single VM accounts.
+For example, we have kept class VM's running and had all students log on to the same VM earlier in the semester, and then progress to setting up their own VM's later in the semester.
+Now we turn our attention to cloud provider options.
 
-We have had generally an excellent experience using the publicly funded cloud, Jetstream2.
-The two big benefits of this model are that it's free for the students, and it's a relatively simple way for the students to get real cloud experience..
-In this model, you apply for an allocation, Jetstream reviews and awards you a certain number of compute hours, you set up the accounts.
-To do the work, each student creates their own virtual machine and uses it to do the data analysis.
-We applied, was approved, and had everything set up in less than a week.
+% TODO: mention RStudio somewhere
 
-The website \url{https://jetstream-cloud.org/index.html} states:
-TODO: cite
+\subsubsection{Public Cloud}
 
+We have had an excellent experience teaching Stat 129 using the NSF funded cloud provider, Jetstream2 \cite{hancock2021jetstream2}.
+In this model, the instructor applies for an allocation, the reviewers award a certain number of compute hours, and the instructor sets up the accounts on the machines.
+We applied, were approved, and had everything set up in less than a week.
+In fact, the reviewer awarded
+
+The Jetstream2 website\footnote{\url{https://jetstream-cloud.org/index.html}} states:
 \begin{quote}
@@
 
-The Jetstream2 user interface and options are far simpler than commercial cloud providers.
+The benefits of Jetstream2 are the free cost for students, the simplicity compared to commercial cloud providers, and the ability to connect seamlessly to large data sets.
+Thanks to NSF funding, our students, many of whom are low income, received accounts without needing to provide a credit card.
+The user interface and options are much simpler than commercial cloud providers, and they cater directly to the academic wishing to analyze large data sets through Manila Filesystems as a service.
+We followed the Jetstream2 documentation and mounted a 1 TB Manila file share to the VMs, which allowed students to simultaneously access and analyze shared data as if it was local to their machine \cite{manila}.
+In our experience, Jetstream2 has lived up to its claims of making cloud computing easy and accessible.
 
-Two disadvantages of Jetstream2 for teaching are the security model, and the lack of safeguards.
-Jetstream2 seems to be focused on allowing individual researchers and small groups, and enabling them to do projects with more compute than they would otherwise be able to.
-What we did, teaching a class where every student gets access, is certainly not the primary use case.
+The disadvantages of Jetstream2 for teaching college classes are the security model, the lack of safeguards, and the annual renewals.
+Jetstream2 caters to the needs of individual researchers and small research groups, and they support this primary use case very well.
+In contrast, teaching a semester long undergraduate class is certainly not Jetstream2's primary use case.
 When you're awarded an allocation, every user gets complete access to every part of the account.
-Students have administrative access to everything the instructor creates, and every other students machine.
+Students have administrative access to everything the instructor creates, and every other student's machine.
 There are no limits on how much compute an individual can use, so it's possible for one user to use up all the credits.
@@
 Even so, a student unintentionally left a large machine running for several weeks.
-We wasn't dilligently monitoring, so this used up around 25\% of our compute credits.
+We weren't monitoring as diligently as we should have, so this used up a significant portion of our 100,000 service units.
 Fortunately, we still had plenty to get through the course.
+Finally, our allocation on Jetstream2 requires annual renewal, which makes us hesitate to rely on it as a long term solution for teaching this class.
+We expect they would renew every year, but there's no guarantee.
 
-A technical detail you'll have to handle is how students will access the large data sets.
-Jetstream2 provides data storage through the Manila Filesystems as a service, and this worked well for sharing large data sets across several different running instances.
-
-Another disadvantage for our use case is the uncertainty of how many years I'll be able to rely on it to continue offering the course, as we need to renew it annually.
-Jetstream2 is not designed as a long term solution for institutions offering the same course year after year. 
-We expect they would renew every year, but there's no guarantee of this happening.
-If they choose not to renew, then I'll need to look for another solution.
-
-\subsection{Commercial Cloud}
+\subsubsection{Commercial Cloud}
 

### 2025-02-04 paper.tex
.

@@
 
-    We compare and contrast different server options for providing undergraduate students with an authentic experience analyzing large, real world data sets located on powerful remote servers.
-These server options include institutionally supported academic computers, commercial cloud services, public cloud services, and personal physical machines.
+% 112 words in abstract
+% 4500 words in document
+    We compare and contrast different compute server options for teaching an undergraduate statistics class centered on analyzing large, real world data sets located on powerful remote servers.
+These options include institutionally supported servers, personal physical servers, public cloud services, and commercial cloud services.
 % TODO: Align these options with the sections
-We briefly describe the course, and offer suggestions on when to choose which server option, based on the background of the instructor and the level of institutional support.
-Looking to the future, we suggest requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having to manage and administer their own server.
+We briefly describe the course, share our experience with each of these options including advantages and disadvantages, mention some technical considerations, and suggest when to choose which option.
+Looking to the future, we outline requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having to manage and administer their own server.
 
@@
 
-\subsection{Dedicated Support}
+\subsection{Institutional Support}
 
@@
 Even so, a student unintentionally left a large machine running for several weeks.
-We weren't monitoring as diligently as we should have, so this used up a significant portion of our 100,000 service units.
-Fortunately, we still had plenty to get through the course.
-Finally, our allocation on Jetstream2 requires annual renewal, which makes us hesitate to rely on it as a long term solution for teaching this class.
+We weren't monitoring as diligently as we should have, and this used up a significant portion of our 100,000 service units, but we still had plenty to get through the course.
+Finally, our allocation on Jetstream2 requires annual renewal, which makes us hesitate to rely on it as a long term solution for teaching Stat 129.
 We expect they would renew every year, but there's no guarantee.
@@
 If it was easier to set up and run a server, then we believe that more instructors would be interested in teaching this kind of class.
-For a lone instructor with no support, the barriers to entry are significant.
+For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
 We would welcome a solution that provides this capability publicly.
@@
 \item \textbf{command line access} If this is present, then it should be possible to install and use all manner of command line software, including R, Python, and SQL.
-\item \textbf{100 GB data storage} 
+\item \textbf{100 GB shared data storage} It should be easy to share the same data with every student.
+    1 TB would be better.
 \item \textbf{free or low cost} Free for students is ideal, of course.
@@
 \item \textbf{longevity} It's much easier for instructors to invest in a particular solution if they can be sure that it will be around in 5 years.
+\item \textbf{load balancing} Current cloud solutions don't do a great job.
 \end{enumerate}

### 2025-02-06 paper.tex
.

@@
 We applied, were approved, and had everything set up in less than a week.
-In fact, the reviewer awarded
+In fact, the reviewer awarded TODO explain SU's
 
@@
 The user interface and options are much simpler than commercial cloud providers, and they cater directly to the academic wishing to analyze large data sets through Manila Filesystems as a service.
+TODO: screenshot of UI
 We followed the Jetstream2 documentation and mounted a 1 TB Manila file share to the VMs, which allowed students to simultaneously access and analyze shared data as if it was local to their machine \cite{manila}.

### 2025-02-11 paper.tex
.

@@
 
-In Stat 129, students process data through a virtual machine (VM), a logical rather than a physical computer.
+The cloud offers the most flexibility of all the server options.
+Indeed, all the different ways one could potentially use and teach cloud services can be overwhelming.
+In Stat 129, students process data by logging into a virtual machine (VM), a logical rather than a physical computer.
 We've experimented with two ways of providing students with cloud access: administrative accounts on the cloud, or local user accounts on a VM.
@@
 The downside here is that you need to pay for a large enough machine all the time, or else manually load balance by switching to smaller or larger machines as the demand changes.
-If we were goint
+%If we were goint
 
-Note that providing students with administrative cloud accounts is not mutually exclusive from simpler single VM accounts.
-For example, we have kept class VM's running and had all students log on to the same VM earlier in the semester, and then progress to setting up their own VM's later in the semester.
+Note that you can combine the simpler single VM accounts together with the richer administrative accounts.
+For example, we have started the semester with a single VM large enough for the class, which students all students log in to and use for simple assignments and in class activities, and then later in the semester students progress to setting up their own VM's.
 Now we turn our attention to cloud provider options.
@@
 The disadvantages of Jetstream2 for teaching college classes are the security model, the lack of safeguards, and the annual renewals.
-Jetstream2 caters to the needs of individual researchers and small research groups, and they support this primary use case very well.
-In contrast, teaching a semester long undergraduate class is certainly not Jetstream2's primary use case.
-When you're awarded an allocation, every user gets complete access to every part of the account.
-Students have administrative access to everything the instructor creates, and every other student's machine.
-There are no limits on how much compute an individual can use, so it's possible for one user to use up all the credits.
+Jetstream2 prioritizes the needs of individual researchers and small research groups, and they support this primary use case very well.
+Jetstream2 is not designed primarily for teaching a semester long undergraduate class with 30 students.
+When the reviewers award the instructor an allocation, every user gets complete access to every part of the account on the allocation.
+If you give students user accounts so that they can have the administrative experience of creating their own VM's, then they will have administrative access to everything the instructor creates, and to every other student's machine.
+There are no limits on how much compute a particular user can use, so it's possible for one user to use up all the credits for the allocation.
 In other words, the only thing that's stopping them from doing something bad is their own awareness and respect for others.
@@
 We weren't monitoring as diligently as we should have, and this used up a significant portion of our 100,000 service units, but we still had plenty to get through the course.
-Finally, our allocation on Jetstream2 requires annual renewal, which makes us hesitate to rely on it as a long term solution for teaching Stat 129.
-We expect they would renew every year, but there's no guarantee.
+Finally, our allocation on Jetstream2 required annual renewal, which makes us hesitate to rely on it as a long term solution for teaching Stat 129.
+We expect the reviewers would renew every year if there is funding available, but there are no guarantees.
+This isn't an issue if someone is paying for commercial cloud services.
 
@@
 Several companies offer commercial cloud services.
-We focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, so that our students can get marketable experience that they can then put on their resume.
+We focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, to provide our students with marketable experience that they can then put on their resume.
 Our favorite feature of AWS is the direct access to Amazon's high quality open data sets through the \texttt{aws s3} command line tools.
@@
 
-The financial aspect is not ideal for students.
+A sustainable funding model for teaching a class like Stat 129 is to make students pay for commercial cloud services, and this doesn't need to be expensive.
+In Spring 2022, we found that students on AWS can get a rich data processing experience for \$20 or less in one semester, which is less than a low cost textbook for a single course.
 These are commercial services, so a student must have a credit card and be prepared to pay for any compute time and storage that they use.
-In Spring 2022, we found that students on AWS can generally get the data processing experience for \$20 or less in one semester.
-If they make a mistake, say using an expensive service and forgetting to shut it off, then they can easily run up an expensive and unexpected bill that can be tough for a college student to cover.
-In the past, AWS offered an educational program with free credits for students.
-The big selling point for me was that students didn't need to even provide a credit card.
-We used this program, but they discontinued it, leaving me searching for other options.
-For a busy faculty member, this is a waste.
+However, if they make a mistake, say using a powerful VM and forgetting to shut it off, then they can easily run up an expensive and unexpected bill that can be tough for a college student to cover.
+In the past, AWS offered an educational program with free credits for students, which didn't even require a credit card.
+We used this program, but they discontinued it, which left us looking for other free options.
 
-Complexity and vendor specificity are also problems when teaching with cloud services.
-Amazon's EC2 (cloud compute server) ``provides a wide selection of over 750 instance types optimized to fit different use cases.''
-There are also hundreds of different services and several world regions to choose from.
-The services often have company specific names, which makes it difficult for an instructor to understand what they are.
+Complexity and vendor specificity are problems when teaching with cloud services.
+Amazon's EC2 (cloud compute server) allows students to create their own VM's, and it ``provides a wide selection of over 750 instance types optimized to fit different use cases.''\cite{aws-ec2}
+There are also hundreds of different services and many world regions to choose from.
+This is an overwhelming amount of choice for a student or even an instructor, and the available options change regularly.
+Furthermore, the services often have company specific names, which makes it difficult for an instructor to understand what they are.
 Indeed, the main purpose of the `AWS Cloud Practitioner' qualification seems to be to understand all of these proprietary names and services.
-This vendor specific education is all well and good if Amazon is providing the training, however, at a public institution our goal is to focus on the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
+As faculty at a public institution, our goal is to focus on the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
 
-While we feel that the commercial cloud services aren't currently the best choice for me, they might be the best choice for you.
-If you already have experience with a particular cloud provider, or you wish to build that experience, then go for it.
+While we feel that the commercial cloud services aren't currently the best choice for teaching Stat 129 at CSU Sacramento, they might be the best choice for you.
+If you already have experience with a particular cloud provider, or you wish to build that experience from the ground up, then this is an opportunity. 
 If your institution has some connection with a cloud provider, then it may be mutually beneficial to leverage those connections.
-For example, the company may provide cloud services for free in exchange for the ability to recruit graduates into that company.
+For example, the company may provide cloud services for free in exchange for the access to recruit graduates into that company.
+If the goal of your course is to focus more on a specific technology such as R, Python, or SQL then you should consider popular vendors such as Posit Cloud (formerly RStudio), Anaconda Cloud, or Snowflake.
+These vendors use commercial cloud services as a foundation, abstract away many of the complicated details of server administration, and provide a familiar web based user interface.
 
-We do believe that viable options can be built using commercial cloud services.
-For example, RStudio, Anaconda, and Snowflake offer cloud services that are essentially a higher level abstraction and user interface catered towards the data science R, Python, or SQL user.
+None of the available cloud services that we've explored or mentioned are ideal for teaching Stat 129.
+We conclude by describing what emph{would} be ideal, based on our experience.
 
@@
 
-If it was easier to set up and run a server, then we believe that more instructors would be interested in teaching this kind of class.
+If it was easier to set up and run a server, then we believe that more instructors would be interested in teaching courses like Stat 129.
 For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
 We would welcome a solution that provides this capability publicly.
-It must have the following features:
+It should have the following features:
 \begin{enumerate}
-\item \textbf{command line access} If this is present, then it should be possible to install and use all manner of command line software, including R, Python, and SQL.
+\item \textbf{shell access} Students should be able 
 \item \textbf{100 GB shared data storage} It should be easy to share the same data with every student.
@@
     If that's not possible, then the cost over the semester should not exceed that of a reasonably priced textbook.
+\item \textbf{long term sustainability} It's much easier for instructors to invest in a particular solution if they can be sure that it will be around in 5 years.
 \end{enumerate}
@@
 \begin{enumerate}
-\item \textbf{longevity} It's much easier for instructors to invest in a particular solution if they can be sure that it will be around in 5 years.
+\item \textbf{administrative access for instructors} If this is present, then it should be possible to install and use all manner of command line software, including R, Python, and SQL.
 \item \textbf{load balancing} Current cloud solutions don't do a great job.
@@
 
-\section{Scratch}
-
-We've taught an upper division statistics ``Big Data'' course at UC Davis and CSU Sacramento and have gained some perspective on what works well for this course.
-Our goals with the course are to get students comfortable with using remote machines and to do realistic analysis of data sets that don’t fit in memory, typically on the order of 100GB or so.
-
-I’ve tried what feels like an excessive number of ways to run the server for the students to get the `server experience':
-Campus supported Linux cluster (SLURM)
-AWS EC2 - student accounts and Jupyter Notebooks
-Google colab
-NSF Jetstream cloud (Access)
-Physical server in the corner of our office (current solution!)
-
 \end{document}

### 2025-02-12 paper.tex
.

@@
 \doublespacing
+
+\usepackage{graphicx}
 
@@
 Looking to the future, we outline requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having to manage and administer their own server.
+
+\emph{keywords: big data, cloud computing, data science, teaching}
 
@@
 \subsection{Which Data?}
+\label{sec:data}
 
@@
 \subsection{Why a Server?}
+\label{sec:server}
 
@@
 If the machine is in your office, then when you launch a big job, the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which is fun.
-A physical machine in our office is our current solution, after having explored and experienced the others described in this article.
+Figure~\ref{fig:server} shows the physical machine in our office that is our current solution, after having explored and experienced the others described in this article.
 We find that it is the most pragmatic long term solution, requiring the least effort on our part.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{server}
+    \caption{A simple server under a desk works just fine for teaching.}
+    \label{fig:server}
+\end{figure}
 
@@
 We completely own and administrator the machine, so we can install whatever we want, whenever we want.
-It's relatively easy and direct to automate certain grading tasks, for example, one can look at \texttt{history} to see every command a student typed during a quiz.
+It's relatively easy and direct to automate certain grading tasks, for example, an instructor can use administrative privileges to look at \texttt{history} to see every command a student typed during a quiz.
 
@@
 The user interface and options are much simpler than commercial cloud providers, and they cater directly to the academic wishing to analyze large data sets through Manila Filesystems as a service.
-TODO: screenshot of UI
+Figure~\ref{fig:jetstream} shows an example of the clean web user interface to create a new instance of a VM.
 We followed the Jetstream2 documentation and mounted a 1 TB Manila file share to the VMs, which allowed students to simultaneously access and analyze shared data as if it was local to their machine \cite{manila}.
 In our experience, Jetstream2 has lived up to its claims of making cloud computing easy and accessible.
+
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=0.5\textwidth]{jetstream}
+    \caption{The Jetstream2 user interface is far simpler and easier to teach compared to commercial cloud providers.}
+    \label{fig:jetstream}
+\end{figure}
 
@@
 \subsubsection{Commercial Cloud}
+\label{sec:comcloud}
 
@@
 
-\section{Future Steps}
+\section{Moving Forward}
 
-If it was easier to set up and run a server, then we believe that more instructors would be interested in teaching courses like Stat 129.
+If it was easier to set up and run a server for teaching undergraduate data analysis classes like Stat 129, then we believe that more instructors would be interested in teaching these courses.
 For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
-We would welcome a solution that provides this capability publicly.
+We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to instructors simultaneously teaching similar courses at other institutions.
 It should have the following features:
 \begin{enumerate}
-\item \textbf{shell access} Students should be able 
-\item \textbf{100 GB shared data storage} It should be easy to share the same data with every student.
-    1 TB would be better.
+\item \textbf{100 GB shared data storage} It should be easy for an entire class to simultaneously read and share the same large data sets.
+    1 TB of storage would be better.
+\item \textbf{data science tools} Students should be able to use common data science programming technogies such as R, Python, and some form of SQL.
+    One way to provide this is to give the instructor administrative privileges to install software.
+    Web based clients using tools such as Jupyter Notebooks certainly broaden the appeal and potential users of the server, but they are not strictly necessary.
 \item \textbf{free or low cost} Free for students is ideal, of course.
-    If that's not possible, then the cost over the semester should not exceed that of a reasonably priced textbook.
-\item \textbf{long term sustainability} It's much easier for instructors to invest in a particular solution if they can be sure that it will be around in 5 years.
+    If free is not possible, then the cost over the semester should not exceed that of a reasonably priced textbook.
+\item \textbf{instructor support}
+    Instructors without a background in systems administration should be able to teach on the server, and receive some level of technical and community support.
+    This is perhaps the hardest requirement, because the best way to provide dedicated support is through paid staff.
+\item \textbf{long term sustainability} 
+    If instructors and institutions are going to invest in creating courses that rely on a server, then they need to have some assurance that the server will continue to exist.
+    A 1 year pilot program might be a useful start, but it's much more compelling to have a guarantee that a server will be available for the next 5 years.
+\item \textbf{minimal specificity}
+    It's ideal if students in a one semester data science class don't need to learn concepts specific to one particular system or vendor, as discussed in Sections~\ref{sec:server} and~\ref{sec:comcloud}.
+\item \textbf{novice user safeguards} It should not be possible for one student to delete or damage course infrastructure, either accidentally or on purpose.
+    Standard Linux user accounts without administrative privileges work fine.
+\item \textbf{parallel capability}
+    The parallel processing capabilities of the server should exceed those of a common laptop to provide a compelling reason to actually use the server.
+    In practice, if there are 10 CPU's available, then students can experience an order of magnitude speedup compared to a serial program, and this is enough to motivate the techniques in Stat 129.
+    An abundance of compute power is definitely not necessary if instructors follow the advice in Section~\ref{sec:data}.
+\item \textbf{shell access} Students should be able to interact with their computer using bash so that they can use data processing pipelines, scripts, and version control.
 \end{enumerate}
-Based on our observations above, there are several other features that would be nice to have, and we list some of them here.
 
-\begin{enumerate}
-\item \textbf{administrative access for instructors} If this is present, then it should be possible to install and use all manner of command line software, including R, Python, and SQL.
-\item \textbf{load balancing} Current cloud solutions don't do a great job.
-\end{enumerate}
+The above instructor level requirements imply other requirements at the system level, and these require serious thought.
+For example, if a shared solution is going to simultaneously support 20 instructors teaching 400 students, then load balancing is a real concern.
+We leave the system level requirements to the professionals.
+
+Fostering a culture of support, mentorship, and growth for teaching unfamiliar but valuable data analysis technologies is more important than the logistics of setting up a server.
+Academia as a whole, and mathematics and statistics department in particular, need to value programming skills and find ways to support instructors who may be teaching new and unfamiliar topics.
+
+Large data sets are here to stay, and we serve our our statistics students well by teaching them how to actually work with large data.
+It's quite uncommon to see a course designed around the student experience on a server, particularly in a mathematics and statistics department.
+We hope that in the future compute servers become to a statistics program somewhat like physical laboratory space is to a chemistry program: a necessity.
+
+%We hope to see more courses like Stat 129 are offered in the future.
 

### 2025-02-19 paper.tex
hand written edits

@@
 These options include institutionally supported servers, personal physical servers, public cloud services, and commercial cloud services.
-% TODO: Align these options with the sections
-We briefly describe the course, share our experience with each of these options including advantages and disadvantages, mention some technical considerations, and suggest when to choose which option.
-Looking to the future, we outline requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having to manage and administer their own server.
+We briefly describe the course, share our experience with each of these options, including advantages and disadvantages, mention some technical considerations, and suggest when to choose which option.
+Looking to the future, we outline requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having managing and administering their own server.
 
-\emph{keywords: big data, cloud computing, data science, teaching}
+keywords: big data, cloud computing, data science education, programming
 
@@
 
-\section{The Course}
+\section{Introduction: Stat 129}
 
@@
 
-Stat 129 uses technologies including bash, Python, and SQL to analyze large data sets, with the intention of exposing students to the common tools in data science.
-Students tend to be highly motivated, because many of them hope to land jobs requiring these skills after graduation.
+Stat 129 uses technologies including bash, Python, and SQL to analyze large data sets, with the intention of exposing students to common tools in data science.
+Students tend to be highly motivated, working towards a job in data science after graduation.
 Stat 129 requires a programming prerequisite, since it is too advanced for a first programming course.
-Statistics majors take Stat 129 after taking `Statistical Computing', an upper division course focused on data analysis using the R programming language, with content similar to Data Science in a Box, R4DS TODO: Cite.
+%Statistics majors take Stat 129 after taking Stat 128, `Statistical Computing', an upper division course focused on data analysis using the R programming language.
 %Ideally, students would have taken two semesters of programming before this course.
 
-The philosophy behind this class is to provide students with real world data sets and ask open ended questions, following in the footsteps of our teachers and mentors, Deborah Nolan and Duncan Temple Lang \cite{nolan2010computing}.
+The philosophy behind Stat 129 is to provide students with real world data sets and ask open ended questions, following in the footsteps of our teachers and mentors, Deborah Nolan and Duncan Temple Lang \cite{nolan2010computing}.
 We have been teaching data science programming technologies in an academic setting since 2015, and we developed Stat 129 and have taught it each year since 2021.
 Feedback from this data forward approach in general, and this class in particular, has been overwhelmingly positive.
-In Stat 129 students ask real questions based on real data that resides on a real server, and for this reason several alumni have reported that this has been the single most practical and valuable class that they have taken in college.
+Several alumni have reported that this has been the single most practical class they have taken in college.
 
@@
 In a typical introductory data analysis class, a student reads a single data file into R or Python and proceeds with the analysis steps \cite{dsbox}.
-This works great if the data fits into memory, but if the data does not fit into memory, then it simply doesn't work.
-Data sets that are larger than memory require entirely different approaches, such as streaming through to filter out a subset of interest.
+This process works great if the data fits into memory, but if the data does not fit into memory, then it doesn't work at all, because the it fails at the reading step.
+Data sets that are larger than memory require entirely different approaches, such as first streaming through to filter out a subset of interest.
 %Once a data set is this large, it's not possible to simply load the whole thing and start working using R or Python.
@@
 Practically speaking, at the time of this writing, a typical student laptop will have between 4 and 32 GB of memory, and data sets that are on the order of 10-100 GB on disk work well to motivate the techniques.
-We assign common data analysis tasks that typically take 1 to 10 minutes if done in serial, and are an order of magnitude faster if done in parallel.
+We assign data analysis tasks that typically take 1 to 10 minutes if done in serial, and are an order of magnitude faster if done in parallel.
 
@@
 %Everything that we do in the course is based on working with a legitimate, unwieldy, data set.
-When teaching Stat 129, we look for data sets that satisfy the criteria below.
+When teaching Stat 129, we look for data sets that satisfy the following criteria.
 \begin{enumerate}
@@
 	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
-\item \textbf{Format} The storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
+\item \textbf{Format} 
+    The data storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
+    Each semester of Stat 129, we work with at least three data sets: a plain text tabular format such as CSV, a nested format such as XML or JSON, and a SQL database.
+    Typically the data will be compressed, with one data set stored across multiple files.
 	Avoid proprietary, novel, or highly specialized data formats.
-    Each semester of Stat 129, we work with data in a plain text tabular format such as CSV or TSV, and nested formats such as XML or JSON.
-\item \textbf{Background Knowledge} A college student should be able to understand what the data actually represents.
-	If it takes longer than a single class period to explain the core idea of what the data captures, then that data is probably too complex.
+\item \textbf{Background Knowledge}
+    A college student should be able to understand what the data actually represents in a single class period.
+    For example, weather data such as Global Historical Climatology Network-Daily Database \cite{menne2012overview} works great, as each row represents a weather observation for a particular date and location.
 \end{enumerate}
-These criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on genomic data.
+The above criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on the specialized formats of genomic data.
 %Most of these principles We've violated at one time or another, and the difficulties became clear later.
@@
 
+It may be possible to preprocess the data set before providing it to students to make it better suit the goals of the course.
 We often find interesting data sets that are close, but not quite what we're looking for.
-In this case, it may be possible to process the data set before providing it to students to make it better suit the goals of the course.
-As a specific example, consider the tax return data on nonprofits provided by the Internal Revenue Service of the United States\cite{irs990}.
+As a specific example, consider the public tax return data on nonprofits provided by the Internal Revenue Service of the United States\cite{irs990}.
 We removed the XML namespace information in these files so that students learning XPATH can write \texttt{"//ActivityOrMissionDesc"} to find the `activity or mission description' for a particular nonprofit rather than the cumbersome \texttt{//*[local-name()='ActivityOrMissionDesc']}.
@@
 
-The large data sets are the heart of Stat 129.
+Large data sets are the heart of Stat 129.
 To properly analyze them we need to provide our students with a compute server.
@@
 Students must access data and run code on remote servers for a class on `Big Data', because that's where big data is typically found: on remote servers.
-There is a qualitative difference between the experience of running code on a laptop with a monitor that's sitting in front of you, and on a machine that you never actually see, that you can only access through \texttt{ssh}, secure shell.
+There is a qualitative difference between running code on a laptop with a monitor that's sitting in front of you, and on a machine that you never actually see, that you can only access through \texttt{ssh}, secure shell.
 Our goal with embracing the shell is to move students beyond the initial unfamiliarity and discomfort that users experience when they first move away from a Graphical User Interface (GUI).
@@
 
-We teach the classic shell bash, because it has proven to be a durable technology.
-Multiple alumni of Stat 129 have reported back that it was the command line skills that led directly to them getting a variety of jobs.
+We teach the classic shell bash, because it has proven to be a durable and nearly universal technology.
+Multiple alumni of Stat 129 have reported back that it was the command line skills in particular that led directly to them getting a variety of jobs.
 In general, students in technology oriented class such as Stat 129 benefit when instructors focus on ubiquitous and proven core technologies, rather than on what happens to be popular this year, because then they come away with skills that they can apply in many different contexts.
-If one hopes to do anything with more advanced tools such as larger compute clusters or frameworks such Hadoop MapReduce \cite{white2012hadoop}, then a foundational knowledge of bash is essential.
+If one hopes to do anything high performance computing (HPC) or frameworks such Hadoop MapReduce \cite{white2012hadoop}, then a foundational working knowledge of bash is essential.
 
 A single server hits the sweet spot for Stat 129; clusters are less ideal, because of the overhead and the abstraction they introduce.
-We've used a cluster with SLURM \cite{yoo2003slurm} job queueing software, common in high performance computing (HPC), for teaching Stat 141C (Big Data and High Performance Computing) at UC Davis in 2019.
 One topic that we cover in Stat 129 is reading the output of \texttt{top/htop} to identify a bottleneck in a shell pipeline that's streaming hundreds of GB of data.
-It's possible to do this with job queueing software where the job is running on a worker node, but it takes either more advanced knowledge, or something specific to that system.
+It's possible to do this with a cluster, but it takes either more advanced knowledge, or tools specific to that system.
 Whenever possible we try to teach the tools that are the most common and widely applicable, to better serve the students in their future careers.
@@
 Which server option is best depends on the goals of the course, the level of institutional support, the number of students enrolled, the background of the students, and the background of the instructor.
-The overarching goal of Stat 129 is to do real things on real data sets, using professional software.
+The overarching goal of Stat 129 is to statistically analyze real data sets, using professional software.
 Students use the server for every assignment and assessment for 15 weeks, so it makes sense to invest some time teaching them skills and tools with steeper learning curves, such as terminal based text editors.
@@
 
-Institutional support for teaching a class like Stat 129 varies widely.
-For us, there were no servers available to use for teaching at the department, college, or university level; however, the college did provide us with startup funds, which funded the experimentation that led to this paper.
+Here are some of details of our specific situation.
+We had no servers available to use for teaching at the department, college, or university level; however, the college did generously provide us with startup funds, which funded the experimentation that led to this paper.
 The Mathematics and Statistics Department at Sacramento State University plans to offer Stat 129 once a year for the foreseeable future, so we need a long term solution for a server.
@@
 What follows are the server options, ordered by our personal preference.
-There's no one size fits all solution, and what's best for you may well differ from our preferences.
+We have yet to discover a perfect solution, and what's best for you may well differ from our preferences.
 
@@
 
+Institutional support for teaching a class like Stat 129 varies widely.
 If you're in the lucky position of having compute infrastructure with dedicated support in your department, college, or university, then this is an ideal option for teaching classes.
@@
 
-While teaching at UC Davis, the system administrator helped us with the following tasks related to the class:
+We've used a cluster with SLURM \cite{yoo2003slurm} job queueing software, common in high performance computing (HPC), for teaching Stat 141C (Big Data and High Performance Computing) at UC Davis in 2019.
+The system administrator helped us with the following tasks related to the class:
 \begin{itemize}
-\item Selecting the right servers to offer to the class.
+\item Selecting the right servers to use for teaching
 \item Creating user accounts from a student roster.
 \item Installing R, Python, along with various packages and libraries of software.
-\item Setting up and configuring an open source database (Postgres) for use by students.
 \item Troubleshooting connectivity and user account issues.
+    For teaching undergraduate courses, timely, helpful support is much more important than physical compute power.
+\item Setting up and configuring an open source Postgres database for use by students.
 \end{itemize}
 Of these tasks, setting up the Postgres database took the most time and expertise, and we were grateful to have the support.
-A system administrator can do many other things, of course.
-For teaching, timely, helpful support is much more important than physical compute power.
+A system administrator can do many other things as well.
 % Just ask!
@@
 
-A personal physical server is essentially the same as a powerful desktop machine, except that you can access it remotely, and it doesn't need to have a monitor.
+A personal physical server is essentially the same as a powerful desktop machine, except that you can access it remotely, and it doesn't need a monitor.
 The machine itself might be in a rack in a server room on campus, or in someone's office.
-If it's in a server rack then you can generally get a machine with more CPU's and memory, better internet connectivity, and it's not cluttering up your space.
+The server rack option is generally better, as the machines typically have more CPU's and memory, better internet connectivity, and they don't clutter up your space.
 If the machine is in your office, then when you launch a big job, the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which is fun.
@@
     \includegraphics[width=0.5\textwidth]{server}
-    \caption{A simple server under a desk works just fine for teaching.}
+    \caption{A simple server under a desk, our current solution, works fine for teaching.}
     \label{fig:server}
@@
 Students have made mistakes and filled up all the disk space, and we needed to diagnose and fix the issue on our own.
-Pedagogically, this mistake followed by the unusable server highlights the importance of not creating unnecessary copies when working with large amounts of data, and it makes students more aware of how much data they're saving.
-The fix is straightforward: just delete the big files that the students accidentally made copies of.
-A more proactive solution to this common problem is to impose a quota on disk usage.
+Pedagogically, this mistake followed by the unusable server highlights the importance of not creating unnecessary copies when working with large amounts of data, and it makes students more aware of exactly what happens if they inadvertantly uncompress and save an excessive amount of data.
+The fix is straightforward: just delete the big files.
+A more proactive solution to this common problem is to impose a quota on disk usage by the home directory.
 We're taking a risk that the server could fail in a more serious way, but it's a calculated risk.
@@
 This machine cost around \$3,000 when purchased in 2022.
-We always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they are expecting.
+We always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they expect.
 In practice it happened occasionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because we had warned them many times and they knew to expect this.
@@
 
-One technical detail to be aware of is that you must provide remote access so that others can log on to the machine.
+For a server in your office, you must provide remote access so that others can log on to the machine.
 Campus IT handled this by providing a DNS hostname.
@@
 Despite that, it provides a legitimate and simple way to get students using a remote server.
-From a student's perspective, there's no difference between \texttt{ssh} to this machine, or to a virtual machine that you own that's running on the cloud.
+From a student's perspective, there's no difference between logging in via \texttt{ssh} to a physical machine, or to a virtual machine running on the cloud.
 
@@
 The cloud offers the most flexibility of all the server options.
-Indeed, all the different ways one could potentially use and teach cloud services can be overwhelming.
+Indeed, all the different possibilities can be overwhelming.
 In Stat 129, students process data by logging into a virtual machine (VM), a logical rather than a physical computer.
-We've experimented with two ways of providing students with cloud access: administrative accounts on the cloud, or local user accounts on a VM.
-We've used both of these options, and they both have a place, depending on the goals of the class.
+We've experimented with two ways of providing students with cloud access: local user accounts for each student on a VM, and administrative cloud accounts for every student.
+Both options have a place, depending on the goals of the class.
 
@@
 They are the administrators to these machines with \texttt{sudo} access, so they can do anything they wish, and any mistakes they make won't affect anyone else.
-Load balancing has the potential to work better in this scenario, as students can create machines that are as large as they like, and delete them when they are done.
-This would be a good option if the course is designed to teach them more about infrastructure, for example, how to implement a full stack web application.
-This rich experience does come with some additional complexity and responsibility, for example, students need to load and save their results as they create and delete machines.
+Load balancing works better in this scenario, as students can create machines that are as large as they like, and delete them when they are done.
+Cloud accounts would be a good option if the course is designed to teach them more about infrastructure, for example, how to implement a full stack web application.
+This rich experience does come with some additional complexity and responsibility, for example, students need to load and save their work as they create and delete machines.
 
-Local user accounts on a single VM allow students to focus more on the data analysis rather than the cloud ideas and infrastructure.
-In this scenario the instructor creates a VM that's large enough to accommodate all of the students simultaneously, and students access through SSH or a web interface.
-The downside here is that you need to pay for a large enough machine all the time, or else manually load balance by switching to smaller or larger machines as the demand changes.
+Local user accounts on a single VM allow students to focus more on the data analysis, rather than the cloud ideas and infrastructure.
+In this scenario, the instructor creates a VM that's large enough to accommodate all of the students simultaneously, and students access through SSH or a web interface.
+The downside here is that you need to pay for a sufficiently large machine all the time, or else manually load balance by switching to smaller or larger machines as the demand changes.
 %If we were goint
 
-Note that you can combine the simpler single VM accounts together with the richer administrative accounts.
-For example, we have started the semester with a single VM large enough for the class, which students all students log in to and use for simple assignments and in class activities, and then later in the semester students progress to setting up their own VM's.
-Now we turn our attention to cloud provider options.
+You can combine the simpler single VM user accounts together with the richer administrative cloud accounts.
+For example, we have started the semester with a single VM large enough for the class, which students all students log in to and use for simple assignments and in class activities, and then students progress to setting up their own VM's later in the semester.
+
+Now we turn our attention to the two cloud provider options: public and commercial.
 
@@
 We applied, were approved, and had everything set up in less than a week.
-In fact, the reviewer awarded TODO explain SU's
+The reviewer awarded $100,000$ service units to teach the class, which was several times more than we requested, and this proved to be plenty.
 
@@
 Thanks to NSF funding, our students, many of whom are low income, received accounts without needing to provide a credit card.
-The user interface and options are much simpler than commercial cloud providers, and they cater directly to the academic wishing to analyze large data sets through Manila Filesystems as a service.
-Figure~\ref{fig:jetstream} shows an example of the clean web user interface to create a new instance of a VM.
+The user interface and options are much simpler than commercial cloud providers as Figure~\ref{fig:jetstream} shows. 
+Jetstream2 caters directly to the academic wishing to analyze large data sets through Manila Filesystems as a service.
 We followed the Jetstream2 documentation and mounted a 1 TB Manila file share to the VMs, which allowed students to simultaneously access and analyze shared data as if it was local to their machine \cite{manila}.
@@
     \includegraphics[width=0.5\textwidth]{jetstream}
-    \caption{The Jetstream2 user interface is far simpler and easier to teach compared to commercial cloud providers.}
+    \caption{This screenshot shows how to create a new instance of a virtual machine.
+The Jetstream2 user interface is far simpler and easier to teach compared to commercial cloud providers.}
     \label{fig:jetstream}
@@
 There are no limits on how much compute a particular user can use, so it's possible for one user to use up all the credits for the allocation.
-In other words, the only thing that's stopping them from doing something bad is their own awareness and respect for others.
+In other words, the only thing that's stopping them from doing something bad is their own awareness and respect for others, which is not ideal for a group of 20-30 novices.
 We mostly got around these issues by setting very clear expectations for behavior, with the understanding that this whole idea is based on trust.
 Even so, a student unintentionally left a large machine running for several weeks.
-We weren't monitoring as diligently as we should have, and this used up a significant portion of our 100,000 service units, but we still had plenty to get through the course.
+We weren't monitoring as diligently as we should have, and this used up a significant portion of our available compute time, but we still had plenty to get through the course.
 Finally, our allocation on Jetstream2 required annual renewal, which makes us hesitate to rely on it as a long term solution for teaching Stat 129.
 We expect the reviewers would renew every year if there is funding available, but there are no guarantees.
-This isn't an issue if someone is paying for commercial cloud services.
+
+Annual renewal isn't an issue if someone is paying for commercial cloud services, and we turn our attention here next.
 
@@
 Several companies offer commercial cloud services.
-We focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, to provide our students with marketable experience that they can then put on their resume.
-Our favorite feature of AWS is the direct access to Amazon's high quality open data sets through the \texttt{aws s3} command line tools.
-From an instructor's perspective, this eliminates the step of downloading the data and making it available to students.
+We focused mainly on Amazon Web Services (AWS), the biggest and most popular vendor, to provide our students with marketable experience for their resumes.
+Our favorite feature of AWS is the direct access to Amazon's high quality open data sets through the \texttt{aws s3} command line tools, which eliminate the step of downloading the data and making it available to students.
 
@@
 These are commercial services, so a student must have a credit card and be prepared to pay for any compute time and storage that they use.
-However, if they make a mistake, say using a powerful VM and forgetting to shut it off, then they can easily run up an expensive and unexpected bill that can be tough for a college student to cover.
+If students make a mistake, say using a powerful VM and forgetting to shut it off, then they can easily run up an expensive and unexpected bill that can be tough for a college student to cover.
 In the past, AWS offered an educational program with free credits for students, which didn't even require a credit card.
@@
 If your institution has some connection with a cloud provider, then it may be mutually beneficial to leverage those connections.
-For example, the company may provide cloud services for free in exchange for the access to recruit graduates into that company.
+For example, the company could provide cloud services for free in exchange for the access to recruit graduates into that company.
+
 If the goal of your course is to focus more on a specific technology such as R, Python, or SQL then you should consider popular vendors such as Posit Cloud (formerly RStudio), Anaconda Cloud, or Snowflake.
 These vendors use commercial cloud services as a foundation, abstract away many of the complicated details of server administration, and provide a familiar web based user interface.
-
-None of the available cloud services that we've explored or mentioned are ideal for teaching Stat 129.
-We conclude by describing what emph{would} be ideal, based on our experience.
+None of the available cloud services that we've explored or mentioned have proven to be ideal for teaching Stat 129.
+We conclude by describing what \emph{would} be ideal, based on our experience.
 
@@
 For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
-We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to instructors simultaneously teaching similar courses at other institutions.
+We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to many instructors simultaneously teaching similar courses at other institutions.
 It should have the following features:
 \begin{enumerate}
-\item \textbf{100 GB shared data storage} It should be easy for an entire class to simultaneously read and share the same large data sets.
-    1 TB of storage would be better.
 \item \textbf{data science tools} Students should be able to use common data science programming technogies such as R, Python, and some form of SQL.
-    One way to provide this is to give the instructor administrative privileges to install software.
-    Web based clients using tools such as Jupyter Notebooks certainly broaden the appeal and potential users of the server, but they are not strictly necessary.
+    %One way to provide this is to give the instructor administrative privileges to install software.
+    Web based tools such as Jupyter Notebooks certainly broaden the appeal and potential users of the server, but they are not strictly necessary.
 \item \textbf{free or low cost} Free for students is ideal, of course.
@@
     The parallel processing capabilities of the server should exceed those of a common laptop to provide a compelling reason to actually use the server.
-    In practice, if there are 10 CPU's available, then students can experience an order of magnitude speedup compared to a serial program, and this is enough to motivate the techniques in Stat 129.
+    In practice, with 10 CPU's students can experience an order of magnitude speedup compared to a serial program, and this is enough to motivate the techniques in Stat 129.
     An abundance of compute power is definitely not necessary if instructors follow the advice in Section~\ref{sec:data}.
+\item \textbf{shared data storage} It should be easy for an entire class to simultaneously read and share the same large data sets.
+    100 GB of storage would suffice, 1 TB of storage would be great, and 10 TB would be more than enough.
 \item \textbf{shell access} Students should be able to interact with their computer using bash so that they can use data processing pipelines, scripts, and version control.
@@
 It's quite uncommon to see a course designed around the student experience on a server, particularly in a mathematics and statistics department.
-We hope that in the future compute servers become to a statistics program somewhat like physical laboratory space is to a chemistry program: a necessity.
+We hope that in the future compute servers become to a data science bachelor's degree program somewhat like physical laboratory space is to a chemistry program: a necessity.
 

### 2025-02-19 paper.tex
.

@@
 
-\usepackage{setspace}
-\doublespacing
+%\usepackage{setspace}
+%\doublespacing
 
@@
 
+\vspace{0.5in}
 keywords: big data, cloud computing, data science education, programming
@@
 \begin{enumerate}
-\item \textbf{Size} The files should be large enough to motivate the computational techniques we're learning, but no larger, as described above.
-\item \textbf{Publicly Available} The data should be readily available for legal public download.
+\item \textbf{size} The files should be large enough to motivate the computational techniques we're learning, but no larger, as described above.
+\item \textbf{publicly available} The data should be readily available for legal public download.
 	Special permissions, non disclosure agreements, and licensing all take time away from teaching.
-\item \textbf{Format} 
+\item \textbf{format} 
     The data storage format should be a common one that students will encounter in data science work, so that students can directly apply what they learn in class.
@@
 	Avoid proprietary, novel, or highly specialized data formats.
-\item \textbf{Background Knowledge}
+\item \textbf{background knowledge}
     A college student should be able to understand what the data actually represents in a single class period.
@@
     \includegraphics[width=0.5\textwidth]{jetstream}
-    \caption{This screenshot shows how to create a new instance of a virtual machine.
-The Jetstream2 user interface is far simpler and easier to teach compared to commercial cloud providers.}
+    \caption{This screenshot shows how to create a new instance of a virtual machine using Jetstream2.
+This user interface is far simpler and easier to teach compared to commercial cloud providers.}
     \label{fig:jetstream}
@@
 We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to many instructors simultaneously teaching similar courses at other institutions.
-It should have the following features:
+It should have the following features (listed alphabetically):
 \begin{enumerate}

### 2025-02-21 paper.tex
.

@@
 
+\usepackage{authblk}
+
 \title{Compute Servers for Teaching Big Data}
-\author{Clark Fitzgerald}
+\author[1]{Clark Fitzgerald}
+\affil[1]{Department of Mathematics and Statistics \protect\\ California State University Sacramento}
 \begin{document}
@@
 
-% 112 words in abstract
-% 4500 words in document
-    We compare and contrast different compute server options for teaching an undergraduate statistics class centered on analyzing large, real world data sets located on powerful remote servers.
-These options include institutionally supported servers, personal physical servers, public cloud services, and commercial cloud services.
-We briefly describe the course, share our experience with each of these options, including advantages and disadvantages, mention some technical considerations, and suggest when to choose which option.
-Looking to the future, we outline requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of having managing and administering their own server.
+% 98 words in abstract
+% ~5000 words in document
+    We compare and contrast a variety of compute server options for teaching an undergraduate statistics class centered on analyzing large, real world data sets.
+These server options include institutionally supported servers, personal physical servers, public cloud services, and commercial cloud services.
+We briefly describe the course, share our experience with each of these options, advantages and disadvantages, some technical considerations, and suggest when to choose which option.
+Looking to the future, we outline requirements for a broadly shared educational compute service that will allow more instructors to teach similar classes without the overhead of administering their own server.
 
-\vspace{0.5in}
-keywords: big data, cloud computing, data science education, programming
+\vspace{0.2in}
+keywords: big data, cloud computing, data science education, statistical computing
 
@@
 
-\emph{
+\begin{quote}
 Statistical analysis of large, complex data sets.
@@
 Students will access data and run code on remote servers.
-}
+\end{quote}
 
@@
 We have been teaching data science programming technologies in an academic setting since 2015, and we developed Stat 129 and have taught it each year since 2021.
-Feedback from this data forward approach in general, and this class in particular, has been overwhelmingly positive.
-Several alumni have reported that this has been the single most practical class they have taken in college.
+Feedback from this data forward approach has been overwhelmingly positive, with several alumni reporting Stat 129 as the most practical class they have taken in college.
 
@@
 For the purposes of Stat 129, `big data' means a data set that is large enough such that one cannot load the entire data set into a laptop's memory at once.
-In a typical introductory data analysis class, a student reads a single data file into R or Python and proceeds with the analysis steps \cite{dsbox}.
-This process works great if the data fits into memory, but if the data does not fit into memory, then it doesn't work at all, because the it fails at the reading step.
+In a typical introductory data analysis class, a student reads a single data file into R or Python and proceeds with the analysis steps \cite{wickham2023r}.
+This process works great if the data fits into memory, but if the data does not fit into memory, then it doesn't work at all, because it fails at the reading step.
 Data sets that are larger than memory require entirely different approaches, such as first streaming through to filter out a subset of interest.
@@
     Each semester of Stat 129, we work with at least three data sets: a plain text tabular format such as CSV, a nested format such as XML or JSON, and a SQL database.
-    Typically the data will be compressed, with one data set stored across multiple files.
+    The data is usually compressed, with one data set stored across multiple files, and this is fine.
 	Avoid proprietary, novel, or highly specialized data formats.
@@
     A college student should be able to understand what the data actually represents in a single class period.
-    For example, weather data such as Global Historical Climatology Network-Daily Database \cite{menne2012overview} works great, as each row represents a weather observation for a particular date and location.
+    For example, weather data such as Global Historical Climatology Network-Daily Database \cite{menne2012overview} works great, as each row represents a weather observation for a specific date and location.
 \end{enumerate}
-The above criteria don't necessarily apply to more domain specific classes, for example, a bioinformatics class studying genomes would do well to focus on the specialized formats of genomic data.
+The above criteria don't necessarily apply to more domain specific classes. 
+For example, a bioinformatics class studying genomes would do well to focus on the specialized formats of genomic data.
 %Most of these principles We've violated at one time or another, and the difficulties became clear later.
@@
 
-It may be possible to preprocess the data set before providing it to students to make it better suit the goals of the course.
+It may be possible to pre-process the data set before providing it to students to make it better suit the goals of the course.
 We often find interesting data sets that are close, but not quite what we're looking for.
-As a specific example, consider the public tax return data on nonprofits provided by the Internal Revenue Service of the United States\cite{irs990}.
-We removed the XML namespace information in these files so that students learning XPATH can write \texttt{"//ActivityOrMissionDesc"} to find the `activity or mission description' for a particular nonprofit rather than the cumbersome \texttt{//*[local-name()='ActivityOrMissionDesc']}.
+Consider the public tax return data on nonprofits provided by the Internal Revenue Service of the United States\cite{irs990}.
+We removed the XML namespace information in these files so that students learning XPATH can write \texttt{"//ActivityOrMissionDesc"} to find the `activity or mission description' for a nonprofit rather than the cumbersome \texttt{//*[local-name()='ActivityOrMissionDesc']}.
 The latter is distracting for students just learning XPATH.
@@
 We teach the classic shell bash, because it has proven to be a durable and nearly universal technology.
-Multiple alumni of Stat 129 have reported back that it was the command line skills in particular that led directly to them getting a variety of jobs.
+Multiple alumni of Stat 129 have reported back that it was specifically the command line skills that led directly to them getting a variety of jobs.
 In general, students in technology oriented class such as Stat 129 benefit when instructors focus on ubiquitous and proven core technologies, rather than on what happens to be popular this year, because then they come away with skills that they can apply in many different contexts.
@@
 Students use the server for every assignment and assessment for 15 weeks, so it makes sense to invest some time teaching them skills and tools with steeper learning curves, such as terminal based text editors.
-If a different class only used a server for two weeks of the semester to examine one particular data set, then it probably makes more sense to use a more approachable web browser based graphical user interface to the server such as Jupyter Notebooks \cite{kluyver2016jupyter}, which will inform the best choice of server options.
+If a different class only used a server for two weeks of the semester to examine a single data set, then it probably makes more sense to use a more approachable web browser based graphical user interface to the server such as Jupyter Notebooks \cite{kluyver2016jupyter}, which will inform the best choice of server options.
 
-Here are some of details of our specific situation.
+Here are some details of our specific situation.
 We had no servers available to use for teaching at the department, college, or university level; however, the college did generously provide us with startup funds, which funded the experimentation that led to this paper.
 The Mathematics and Statistics Department at Sacramento State University plans to offer Stat 129 once a year for the foreseeable future, so we need a long term solution for a server.
-In a particular year we'll have one instructor, 15-30 students, and no staff, teaching assistants, or other support.
+In a given year we'll have one instructor, 15-30 students, and no staff, teaching assistants, or other support.
 We don't need a large machine for any purpose other than teaching.
-We have some professional background in programming, and our server administration experience has proven particularly useful.
+We have some professional background in programming, and our server administration experience has proven especially useful.
 
@@
 \begin{itemize}
-\item Selecting the right servers to use for teaching
+\item Selecting the right servers to use for teaching.
 \item Creating user accounts from a student roster.
@@
 A system administrator can do many other things as well.
-% Just ask!
+If you don't have institutional support, then it's possible to be your own systems administrator, and we consider this situation next.
 
@@
 The machine itself might be in a rack in a server room on campus, or in someone's office.
-The server rack option is generally better, as the machines typically have more CPU's and memory, better internet connectivity, and they don't clutter up your space.
-If the machine is in your office, then when you launch a big job, the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which is fun.
+The server rack option is generally preferable, as the machines have more CPU's and memory, better internet connectivity, and they don't clutter up your space.
 Figure~\ref{fig:server} shows the physical machine in our office that is our current solution, after having explored and experienced the others described in this article.
@@
 The major advantages of the personal physical server are cost and simplicity.
-Regarding cost, we've found it easier to fund a one time purchase of a few thousand dollars compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
+Regarding cost, we've found it easier to fund a one time purchase of a few thousand dollars, compared to finding a sponsor who will commit to the ongoing variable subscription fees of the cloud services, even if the total cost would be lower for the cloud subscription.
 It's simple because you can set it up one time, and then use the same machine for several years.
@@
 It's relatively easy and direct to automate certain grading tasks, for example, an instructor can use administrative privileges to look at \texttt{history} to see every command a student typed during a quiz.
+If the machine is in your office, then when you launch a big job, the machine will heat up and the cooling fan will turn on, which provides a very real connection to the amount of data being processed for the students in office hours, which is fun.
 
@@
 
-The server doesn't need to be particularly powerful or expensive.
+The server doesn't need to be powerful or expensive.
 For a class of 15 students, we found that a single 32 core CPU, a GPU, and 128 GB of memory was just sufficient for students to experience the class assignments in the way we intended.
@@
 We always recommend that students do their homework well before the due date, because if a handful of users are simultaneously trying to use parallel jobs with all available CPU cores, then they won't see the speedup they expect.
-In practice it happened occasionaly that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because we had warned them many times and they knew to expect this.
+In practice, it happened occasionally that students who waited until the last hour to do an assignment didn't get to see their final result, but they weren't surprised, because we had warned them many times and they knew to expect this.
 We've used a 16 core CPU machine in previous semesters, and found that it was not sufficient for 15 students.
@@
 For a server in your office, you must provide remote access so that others can log on to the machine.
-Campus IT handled this by providing a DNS hostname.
-Our students need to be either on campus or at home on the campus VPN (Virtual Private Network), and they can login via ssh, for example: \texttt{ssh fitzgerald@nsm-stats.csus.edu}.
+Campus IT handled this by providing a DNS host name.
+Our students need to be either on campus or at home on the campus VPN (Virtual Private Network), and they can login via \texttt{ssh}.
 We recommend against setting up a physical server at your home, because providing reliable remote access could be tough depending on your internet provider, and there's no reason you should pay for the electricity instead of campus.
@@
 Local user accounts on a single VM allow students to focus more on the data analysis, rather than the cloud ideas and infrastructure.
-In this scenario, the instructor creates a VM that's large enough to accommodate all of the students simultaneously, and students access through SSH or a web interface.
+In this scenario, the instructor creates a VM that's large enough to accommodate all of the students simultaneously, and students access through \texttt{ssh} or a web interface.
 The downside here is that you need to pay for a sufficiently large machine all the time, or else manually load balance by switching to smaller or larger machines as the demand changes.
@@
 We applied, were approved, and had everything set up in less than a week.
-The reviewer awarded $100,000$ service units to teach the class, which was several times more than we requested, and this proved to be plenty.
+The reviewer awarded $100,000$ service units to teach the class, which was several times more than we requested, and this proved to be plenty for 20 students over a semester.
 
@@
 
-The disadvantages of Jetstream2 for teaching college classes are the security model, the lack of safeguards, and the annual renewals.
+The main disadvantages of Jetstream2 for teaching college classes are the security model and the lack of safeguards.
 Jetstream2 prioritizes the needs of individual researchers and small research groups, and they support this primary use case very well.
-Jetstream2 is not designed primarily for teaching a semester long undergraduate class with 30 students.
-When the reviewers award the instructor an allocation, every user gets complete access to every part of the account on the allocation.
-If you give students user accounts so that they can have the administrative experience of creating their own VM's, then they will have administrative access to everything the instructor creates, and to every other student's machine.
-There are no limits on how much compute a particular user can use, so it's possible for one user to use up all the credits for the allocation.
+Every user gets complete access to every part of the account on the allocation, which keeps things simple.
+However, if you give students user accounts so that they can have the experience of creating their own VM's, then they will have administrative access to everything the instructor creates, and to every other student's machine.
+There are no limits on how much compute a user can use, so it's possible for one user to use up all the credits for the allocation.
 In other words, the only thing that's stopping them from doing something bad is their own awareness and respect for others, which is not ideal for a group of 20-30 novices.
-We mostly got around these issues by setting very clear expectations for behavior, with the understanding that this whole idea is based on trust.
+We mostly got around these issues by setting very clear expectations for behavior, with the understanding that this whole class is based on trust.
 Even so, a student unintentionally left a large machine running for several weeks.
 We weren't monitoring as diligently as we should have, and this used up a significant portion of our available compute time, but we still had plenty to get through the course.
+
 Finally, our allocation on Jetstream2 required annual renewal, which makes us hesitate to rely on it as a long term solution for teaching Stat 129.
 We expect the reviewers would renew every year if there is funding available, but there are no guarantees.
-
 Annual renewal isn't an issue if someone is paying for commercial cloud services, and we turn our attention here next.
@@
 Indeed, the main purpose of the `AWS Cloud Practitioner' qualification seems to be to understand all of these proprietary names and services.
-As faculty at a public institution, our goal is to focus on the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
+As faculty at a public institution, our goal is to teach the overarching concepts and portable technologies that a student can take and apply anywhere, regardless of the specific vendor.
 
@@
 
-If the goal of your course is to focus more on a specific technology such as R, Python, or SQL then you should consider popular vendors such as Posit Cloud (formerly RStudio), Anaconda Cloud, or Snowflake.
+If the goal of your course is to focus more on a specific technology such as R, Python, or SQL, then you should consider popular vendors such as Posit Cloud (formerly RStudio), Anaconda Cloud, or Snowflake.
 These vendors use commercial cloud services as a foundation, abstract away many of the complicated details of server administration, and provide a familiar web based user interface.
@@
 
-If it was easier to set up and run a server for teaching undergraduate data analysis classes like Stat 129, then we believe that more instructors would be interested in teaching these courses.
+There are many technical details and considerations when relying on a server to teach an undergraduate data analysis class like Stat 129, and we've only provided a broad overview here.
 For a lone instructor with a heavy teaching load and no support, the barriers to entry are significant.
-We would welcome a shared solution, cloud or otherwise, that provides this capability publicly to many instructors simultaneously teaching similar courses at other institutions.
+If it was easier to procure, configure, and run a server, then we believe that more instructors would be interested in teaching these courses.
+
+We would welcome a shared solution, cloud based or otherwise, that provides this capability publicly to many instructors simultaneously teaching similar courses at other institutions.
 It should have the following features (listed alphabetically):
 \begin{enumerate}
-\item \textbf{data science tools} Students should be able to use common data science programming technogies such as R, Python, and some form of SQL.
+\item \textbf{data science tools} Students should be able to use common data science programming technologies such as R, Python, and some form of SQL.
     %One way to provide this is to give the instructor administrative privileges to install software.
-    Web based tools such as Jupyter Notebooks certainly broaden the appeal and potential users of the server, but they are not strictly necessary.
+    Web based tools such as Jupyter Notebooks certainly broaden the appeal and potential user base of the server, but they are not strictly necessary.
 \item \textbf{free or low cost} Free for students is ideal, of course.
@@
 \item \textbf{minimal specificity}
-    It's ideal if students in a one semester data science class don't need to learn concepts specific to one particular system or vendor, as discussed in Sections~\ref{sec:server} and~\ref{sec:comcloud}.
+    It's ideal if students in a one semester data science class don't need to learn concepts specific to one system or vendor, as discussed in Sections~\ref{sec:server} and~\ref{sec:comcloud}.
 \item \textbf{novice user safeguards} It should not be possible for one student to delete or damage course infrastructure, either accidentally or on purpose.
@@
 
-Fostering a culture of support, mentorship, and growth for teaching unfamiliar but valuable data analysis technologies is more important than the logistics of setting up a server.
-Academia as a whole, and mathematics and statistics department in particular, need to value programming skills and find ways to support instructors who may be teaching new and unfamiliar topics.
+Fostering a culture of support, mentorship, and growth for instructors is more important than the logistics of setting up a server.
+%It's hard to integrate a server experience throughout an undergraduate class, especially compared to a traditional mathematics or statistics course,.
+We academics should value programming skills, and find ways to support instructors who may be teaching new and unfamiliar topics.
+Some specific things we can do are: encourage our peers to accept the challenge of teaching material that may be outside of their comfort zones, share our course materials, and meet with peers regularly when they're teaching this new material for the first time.
 
-Large data sets are here to stay, and we serve our our statistics students well by teaching them how to actually work with large data.
+Large data sets are here to stay, and we serve our students well by teaching them how to actually work with large data.
 It's quite uncommon to see a course designed around the student experience on a server, particularly in a mathematics and statistics department.
-We hope that in the future compute servers become to a data science bachelor's degree program somewhat like physical laboratory space is to a chemistry program: a necessity.
+We hope that, in the future, compute servers become to a data science bachelor's degree program somewhat like physical laboratory space is to a chemistry program: a necessity.
 
